---
title: "Gilmore Girls coffee addiction analysis with R"
author: "Dr. Shirin Glander"
date: '`r Sys.Date()`'
output: html_document
---

Last week's showed how to create a [character network with Gilmore Girls transcripts](https://shiring.github.io/text_analysis/2016/11/13/gilmore_girls_part1_post).

This week I want to explore the Gilmore Girls famous coffee addiction by analysing the same transcripts.

I am also showcasing how to use the [recently updated ggplot2 2.2.0](https://blog.rstudio.org/2016/11/14/ggplot2-2-2-0/).

The transcripts were prepared as described in last week's post.

```{r echo = FALSE, message = FALSE, warning = FALSE, cache=FALSE}
transcripts <- read.table("gilmore_girls_transcripts.txt", header = TRUE)

transcripts$thepage <- as.character(transcripts$thepage)  # convert to character vector
transcripts <- transcripts[!transcripts$thepage == "", ]  # remove empty lines

# separate first column after first colon
library(tidyr)
transcripts_2 <- separate(transcripts, "thepage", into = c("character", "dialogue"), sep = ":", extra = "merge", fill = "right")

# remove leading and trailing whitespace
transcripts_2$character <- gsub("^\\s+|\\s+$", "", transcripts_2$character)

# convert all character names to all upper case
transcripts_2$character <- toupper(transcripts_2$character)

# fix misspelled character names
transcripts_2$character <- gsub("ZACK", "ZACH", transcripts_2$character)
transcripts_2$character <- gsub("LORLEAI", "LORELAI", transcripts_2$character)
transcripts_2$character <- gsub("LOREALI", "LORELAI", transcripts_2$character)
transcripts_2$character <- gsub("LORELI", "LORELAI", transcripts_2$character)
transcripts_2$character <- gsub("LORLAI", "LORELAI", transcripts_2$character)
transcripts_2$character <- gsub("LORELA$", "LORELAI", transcripts_2$character)
transcripts_2$character <- gsub("LORLELAI", "LORELAI", transcripts_2$character)
transcripts_2$character <- gsub("^ORELAI", "LORELAI", transcripts_2$character)
transcripts_2$character <- gsub("LOREAI", "LORELAI", transcripts_2$character)
transcripts_2$character <- gsub("^ORY", "RORY", transcripts_2$character)
transcripts_2$character <- gsub("LUK$", "LUKE", transcripts_2$character)
transcripts_2$character <- gsub("BABETE", "BABETTE", transcripts_2$character)
transcripts_2$character <- gsub("BABETTER", "BABETTE", transcripts_2$character)
transcripts_2$character <- gsub("BARBETTE", "BABETTE", transcripts_2$character)
transcripts_2$character <- gsub("BABETTE/MISS PATTY", "BABETTE AND MISS PATTY", transcripts_2$character)
transcripts_2$character <- gsub("JACKSON/SOOKIE", "JACKSON AND SOOKIE", transcripts_2$character)
transcripts_2$character <- gsub("LORELAI/SOOKIE", "LORELAI AND SOOKIE", transcripts_2$character)
transcripts_2$character <- gsub("LORELAI/RORY", "LORELAI AND RORY", transcripts_2$character)
transcripts_2$character <- gsub("TAYOR", "TAYLOR", transcripts_2$character)
transcripts_2$character <- gsub("TRISTIN", "TRISTAN", transcripts_2$character)
transcripts_2$character <- gsub("MICHE$", "MICHEL", transcripts_2$character)
transcripts_2$character <- gsub("MICHELL", "MICHEL", transcripts_2$character)
transcripts_2$character <- gsub("SOOKI$", "SOOKIE", transcripts_2$character)
transcripts_2$character <- gsub("SOOKEI", "SOOKIE", transcripts_2$character)
transcripts_2$character <- gsub("SOOKIES", "SOOKIE", transcripts_2$character)
transcripts_2$character <- gsub("Mrs.KIM", "MRS KIM", transcripts_2$character)
transcripts_2$character <- gsub("MRS.KIM", "MRS KIM", transcripts_2$character)
transcripts_2$character <- gsub("MRS KIM", "MRS KIM", transcripts_2$character)
transcripts_2$character <- gsub("RICHRAD", "RICHARD", transcripts_2$character)
transcripts_2$character <- gsub("RMILY", "EMILY", transcripts_2$character)
transcripts_2$character <- gsub("CHRISTOHPER", "CHRISTOPHER", transcripts_2$character)
transcripts_2$character <- gsub("CHRISTOPER", "CHRISTOPHER", transcripts_2$character)
transcripts_2$character <- gsub("CHRSTOPHER", "CHRISTOPHER", transcripts_2$character)
transcripts_2$character <- gsub("CHRIS$", "CHRISTOPHER", transcripts_2$character)
transcripts_2$character <- gsub("CHERRY", "SHERRY", transcripts_2$character)
transcripts_2$character <- gsub("LINDAY", "LINDSAY", transcripts_2$character)

# substitute &#146; with apostrophe
transcripts_2$character <- gsub("&#146;", "'", transcripts_2$character)

# some ANDs are written as &AMP; so they will be changed as well
transcripts_2$character <- gsub("&AMP;", "AND", transcripts_2$character)

# and finally I want ANDs to be written as semicolons
transcripts_2$character <- gsub(" AND ", ";", transcripts_2$character)

#  and remove disclaimer lines
transcripts_2 <- transcripts_2[-which(transcripts_2$character == "DISCLAIMER"), ] 

library(splitstackshape)
transcripts_2 <- cSplit(transcripts_2, splitCols = "character", sep = ";", direction = "long")

# separating all rows where multiple characters spoke into one line per character with duplicate line text
library(splitstackshape)
transcripts_2 <- cSplit(transcripts_2, splitCols = "character", sep = ";", direction = "long")

transcripts_2$dialogue <- gsub("&#146;", "'", transcripts_2$dialogue)
```

<br>

## How often do they talk about coffee?

The first question I want to explore is how often Lorelai, Rory, Luke, Sookie, Lane and Paris talk about coffee. For this, I am subsetting the lines of these characters to those, in which they talk about coffee (i.e. that include the word "coffee" at least once). I then created a plot showing how often they talk about coffee per episode and season.

```{r echo = TRUE, message = FALSE, warning = FALSE, cache=FALSE}
# set names
names <- c("LORELAI", "RORY", "LUKE", "SOOKIE", "LANE", "PARIS")

# extract lines for each character in names
for (name in names){
  assign(paste("lines", name, sep = "_"), transcripts_2[grep(paste0(name), transcripts_2$character), ])
}

# extract lines not from characters in names
lines_OTHER <- transcripts_2[!grepl("LORELAI|RORY|LUKE|SOOKIE|LANE|PARIS", transcripts_2$character), ]

# update names to include "other"
names <- c(names, "OTHER")
```

```{r echo = TRUE, message = FALSE, warning = FALSE, cache=FALSE}
# which lines mention coffee.
for (name in names){
  df <- get(paste("lines", name, sep = "_"))
  assign(paste("coffee", name, sep = "_"), df[grep("coffee", df$dialogue), ])
}
```

```{r echo = TRUE, message = FALSE, warning = FALSE, cache=FALSE}
# calculate number of lines in total and with coffee mentions per episode
library(dplyr)

for (name in names){
  lines_p_ep <- get(paste("lines", name, sep = "_"))
  lines_p_ep <- as.data.frame(table(lines_p_ep$episode))
  
  df <- get(paste("coffee", name, sep = "_"))
  df_2 <- as.data.frame(table(df$episode))
  
  # calculate coffee lines per episode
  coffee_lines_p_ep <- full_join(lines_p_ep, df_2, by = "Var1")
  coffee_lines_p_ep$ratio_p_ep <- coffee_lines_p_ep$Freq.y/coffee_lines_p_ep$Freq.x
  
  coffee_lines_p_ep$Season <- gsub("_.*", "", coffee_lines_p_ep$Var1)
  coffee_lines_p_ep$character <- paste(name)
  
  assign(paste("coffee_lines_p_ep", name, sep = "_"), coffee_lines_p_ep)
}

coffee_df <- rbind(coffee_lines_p_ep_LORELAI, coffee_lines_p_ep_RORY, coffee_lines_p_ep_LUKE, coffee_lines_p_ep_SOOKIE, 
                   coffee_lines_p_ep_LANE, coffee_lines_p_ep_PARIS)
```

```{r echo = TRUE, message = FALSE, warning = FALSE, cache=FALSE}
library(ggplot2)

my_theme <- function(base_size = 12, base_family = "sans"){
  theme_grey(base_size = base_size, base_family = base_family) +
  theme(
    axis.text = element_text(size = 12),
    axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5),
    axis.title = element_text(size = 14),
    panel.grid.major = element_line(color = "grey"),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "aliceblue"),
    strip.background = element_rect(fill = "lightgrey", color = "grey", size = 1),
    strip.text = element_text(face = "bold", size = 12, color = "black"),
    legend.position = "bottom",
    legend.justification = "top", 
    legend.box = "horizontal",
    legend.box.background = element_rect(colour = "grey50"),
    legend.background = element_blank(),
    panel.border = element_rect(color = "grey", fill = NA, size = 0.5)
  )
}
```

```{r echo = TRUE, message = FALSE, warning = FALSE, cache=FALSE, fig.width = 14, fig.height = 7, fig.align = "center"}
library(tidyr)

coffee_df_gather <- coffee_df %>% 
  gather(group, value, Freq.x:ratio_p_ep)

coffee_df_gather[which(coffee_df_gather$group == "Freq.x"), "group"] <- "Lines per episode"
coffee_df_gather[which(coffee_df_gather$group == "Freq.y"), "group"] <- "Coffee lines per episode"
coffee_df_gather[which(coffee_df_gather$group == "ratio_p_ep"), "group"] <- "Ratio of coffee lines per episode"

coffee_df_gather$character <- factor(coffee_df_gather$character, levels = names)
coffee_df_gather$group <- factor(coffee_df_gather$group, levels = c("Lines per episode", "Coffee lines per episode", "Ratio of coffee lines per episode"))

ggplot(data = coffee_df_gather, aes(x = character, y = value, color = Season, fill = Season)) +
  geom_boxplot(alpha = 0.5) +
  labs(
    x = "",
    y = "Number or ratio of lines",
    title = "Lorelai - the coffee queen",
    subtitle = "Coffee mentions per episode and season",
    caption = "\nThese boxplots show the number of total lines per episode and season, lines with coffee mentions per epsiode and season,
     as well as the ratio between the two for the main characters Lorelai, Rory, Luke, Sookie, Lane and Paris. Lorelai consistently had
     the most lines per episode, followed closely by Rory and Luke. Sookie, Lane and Paris had roughly the same numbers of lines per episode,
     although their variance is higher. The same trend is reflected by the number of lines with coffee mentions, except that Rory and Luke seem
     to talk similarly often about coffee. Interestingly, Lorelai's (verbal) coffee obsession seems to have decreased slightly over the seasons.
     The ratio of coffee lines divided by the total number of lines for each episode reflects that even though Lorelai, Rory and Luke talk a lot about coffee,
     they also talk a lot in general, so their ratio of coffee mentions is only slightly higher than Sookie's, Lane's and Paris'. 
     While the latter's mean ratio is lower. they have a much higher variance with a few episodes where they seem to have talked relatively much about coffee.
     During the first seasons, Luke is the character with the highest ratio of coffee vs general talk but this is decreasing over time and becomes more
     similar to Lorelai and Rory's coffee talk ratio."
  ) +
  my_theme() +
  facet_wrap(~ group, ncol = 3, scales = "free") +
  guides(fill = guide_legend(nrow = 1, byrow = TRUE)) +
  scale_fill_brewer(palette="Dark2") +
  scale_color_brewer(palette="Dark2")
```

<br>

The differences we seen in coffee mentions between characters are of course somewhat biased by differences in the total number of lines of each character in the respective episodes, as well by the length of the episodes.

```{r echo = TRUE, message = FALSE, warning = FALSE, cache=FALSE}
# calculate number of lines in total
for (name in names){
  lines_p_ep <- get(paste("lines", name, sep = "_"))
  lines_p_ep <- as.data.frame(table(lines_p_ep_pre$episode))

  lines_p_ep$character <- paste(name)
  
  assign(paste("lines_p_ep", name, sep = "_"), lines_p_ep)
}

lines_df_wide <- spread(rbind(lines_p_ep_LORELAI, lines_p_ep_RORY, lines_p_ep_LUKE, lines_p_ep_SOOKIE, 
                              lines_p_ep_LANE, lines_p_ep_PARIS, lines_p_ep_OTHER), Var1, Freq)
rownames(lines_df_wide) <- lines_df_wide[, 1]
lines_df_wide <- as.data.frame(t(lines_df_wide[, -1]))

prop_lines_df_wide <- sweep(lines_df_wide, 1, rowSums(lines_df_wide), FUN="/")
percent_lines_df_wide <- prop_lines_df_wide * 100
percent_lines_df_wide$Episode <- rownames(percent_lines_df_wide)
percent_lines_df_wide$Season <- gsub("_.*", "", percent_lines_df_wide$Episode)
                      
percent_lines_df_gather <- percent_lines_df_wide %>% 
  gather(Character, Percent, LANE:SOOKIE)
```

```{r echo = TRUE, message = FALSE, warning = FALSE, cache=FALSE, fig.width = 14, fig.height = 23, fig.align = "center"}
percent_lines_df_gather$Character <- factor(percent_lines_df_gather$Character, levels = names)
percent_lines_df_gather <- merge(percent_lines_df_gather, subset(lines_LORELAI, select = c(episode, episode_running_nr)), by.x = "Episode", by.y = "episode")

# plot
bp <- ggplot(percent_lines_df_gather, aes(x = "", y = Percent, fill = Character)) + 
  geom_bar(width = 1, stat = "identity") + theme_minimal() +
  scale_fill_brewer(palette="Spectral")

pie <- bp + coord_polar("y", start = 0) +
  theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  panel.border = element_blank(),
  panel.grid = element_blank(),
  axis.ticks = element_blank(),
  plot.title = element_text(size = 14, face = "bold"),
  legend.title = element_blank(),
  legend.position = "top",
  legend.text = element_text(size = 8),
  legend.justification = "top", 
    legend.box = "horizontal",
    legend.box.background = element_rect(colour = "grey50"),
    legend.background = element_blank(),
  ) + guides(fill = guide_legend(nrow = 1, byrow = TRUE))
pie + facet_wrap( ~ Episode, ncol = 11) +
  labs(
    title = "Who's the biggest talker?",
    subtitle = "Proportion of lines per episode for main characters")
```

---

<br>

```{r echo = TRUE, message = FALSE, warning = FALSE, cache=FALSE}
library(tm)
data("crude")
library(SnowballC)

transform_to_space <- content_transformer(function(x, pattern) gsub(pattern, " ", x))

for (name in names){
  lines <- get(paste("lines", name, sep = "_"))$dialogue
  
  corpus <- Corpus(VectorSource(lines))

  # Removing punctuation
  corpus <- tm_map(corpus, transform_to_space, "-")
  corpus <- tm_map(corpus, removePunctuation)
  
  # Transforming to lower case
  corpus <- tm_map(corpus, content_transformer(tolower))
    
  # Stripping numbers
  corpus <- tm_map(corpus, removeNumbers)
    
  # Removing stopwords
  corpus <- tm_map(corpus, removeWords, stopwords("english"))
  
  # Stripping whitespace
  corpus <- tm_map(corpus, stripWhitespace)
  
  # Assigning to object and treat as text document
  assign(paste("corpus", name, sep = "_"), tm_map(corpus, PlainTextDocument))
}

corpus_LORELAI
corpus_RORY
```

```{r echo = TRUE, message = FALSE, warning = FALSE, cache=FALSE}
for (name in names){
  assign(paste("dtm", name, sep = "_"), DocumentTermMatrix(get(paste("corpus", name, sep = "_"))))
}

dtm_LORELAI
dtm_RORY
```

```{r echo = TRUE, message = FALSE, warning = FALSE, cache=FALSE}
# get word frequencies from DTM
for (name in names){
  assign(paste("freq_dtm", name, sep = "_"), colSums(as.matrix(get(paste("dtm", name, sep = "_")))))
}

head(sort(freq_dtm_LORELAI, decreasing = TRUE))
head(sort(freq_dtm_RORY, decreasing = TRUE))
```

```{r echo = TRUE, message = FALSE, warning = FALSE, cache=FALSE}
# find strongest associations with "coffee"
for (name in names){
  assign(paste("coffee_assoc", name, sep = "_"), findAssocs(get(paste("dtm", name, sep = "_")), "coffee", 0))
}

head(sort(coffee_assoc_LORELAI$coffee, decreasing = TRUE))
head(sort(coffee_assoc_RORY$coffee, decreasing = TRUE))
```

<br>

```{r echo = TRUE, message = FALSE, warning = FALSE, cache=FALSE}
dtm_LORELAI_df<- as.data.frame(as.matrix(dtm_LORELAI))
dtm_RORY_df <- as.data.frame(as.matrix(dtm_RORY))

area1 = length(unique(colnames(dtm_LORELAI_df)))
area2 = length(unique(colnames(dtm_RORY_df)))
cross.area = length(unique(which(colnames(dtm_LORELAI_df) %in% colnames(dtm_RORY_df))))

library(VennDiagram)
venn.plot <- draw.pairwise.venn(area1, area2, cross.area,
                   category = c("LORELAI", "RORY"),
                   euler.d = TRUE, 
                   scaled = TRUE, 
                   inverted = FALSE,
                   lwd = rep(2, 2),
                   lty = rep(2, 2), 
                   col = rep("black", 2),
                   fill = c("red", "lightblue"), 
                   alpha = c(0.25, 0.5), 
                   label.col = rep("black", 3), 
                   cex = rep(1, 3), 
                   fontface = rep("plain", 3), 
                   fontfamily = rep("serif", 3),
                   cat.cex = rep(0.8, 2))
grid.draw(venn.plot);
grid.newpage();
```


<br>

-----------------

```{r echo = TRUE, message = FALSE, warning = FALSE, cache=FALSE}
sessionInfo()
```
