---
title: "A simple 'How to map your Google location history'"
author: "Dr. Shirin Glander"
date: '`r Sys.Date()`'
output: html_document
---

It is no secret that Google basically Big Brothers most of us. But at least they allow us to access quite a lot of the data they have collected on us. One of these is the Google location history. If you want to see a few options of how to quickly and easily visualize your location history, stay tuned...

The Google location history can be downloaded from your Google account under [http://takeout.google.com/](http://takeout.google.com/). Make sure you only tick *"location history"* for download, otherwise it will pobably take super long if you ask to get all your Google data.

It is saved as a .json file and can be loaded with the [jsonlite](https://cran.r-project.org/web/packages/jsonlite/index.html) package. Loading this file into R might talk a few minutes because it can be quite big, depending on how much location data Google had saved about you.

```{r echo = TRUE, message = FALSE, warning = FALSE, cache=FALSE}
library(jsonlite)
#system.time(x <- fromJSON("U:/Github_blog/Takeout/Standortverlauf/Standortverlauf.json"))
system.time(x <- fromJSON("~/Documents/Github/Standortverlauf.json"))
```

<br>

The date and time column is in POSIX milliseconds format, so I converted it to the human readable normal POSIX format.

Similarly, longitude and latitude are saved in E7 format and was converted to GPS coordinates.

```{r echo = TRUE, message = FALSE, warning = FALSE, cache=FALSE}
# extracting the locations dataframe
loc = x$locations

# converting time column from posix milliseconds into a readable time scale
loc$time = as.POSIXct(as.numeric(x$locations$timestampMs)/1000, origin = "1970-01-01")

# converting longitude and latitude from E7 to GPS coordinates
loc$lat = loc$latitudeE7 / 1e7
loc$lon = loc$longitudeE7 / 1e7
```

<br>

This is how the data looks like now:

```{r echo = TRUE, message = FALSE, warning = FALSE, cache=FALSE}
head(loc)
```

<br>

# How many data points did it record?

Before we get to actual plotting of map data, I want to explore a few basic statistics of the data about me.

First, how many data points were recorded over which period of time?

```{r echo = TRUE, message = FALSE, warning = FALSE, cache=FALSE}
nrow(loc)
```

```{r echo = TRUE, message = FALSE, warning = FALSE, cache=FALSE}
summary(loc$time)
```

<br>

## And how are they distributed over days, months, years?

```{r echo = TRUE, message = FALSE, warning = FALSE, cache=FALSE}
# calculate the number of data points per day, month and year
library(lubridate)
library(zoo)

loc$date <- as.Date(loc$time, '%Y/%m/%d')
loc$year <- year(loc$date)
loc$month_year <- as.yearmon(loc$date)

points_p_day <- data.frame(table(loc$date), group = "day")
points_p_month <- data.frame(table(loc$month_year), group = "month")
points_p_year <- data.frame(table(loc$year), group = "year")

nrow(points_p_day)
nrow(points_p_month)
nrow(points_p_year)
```

```{r echo = TRUE, message = FALSE, warning = FALSE, cache=FALSE}
# set up plotting theme
library(ggplot2)
library(ggmap)

my_theme <- function(base_size = 12, base_family = "sans"){
  theme_grey(base_size = base_size, base_family = base_family) +
  theme(
    axis.text = element_text(size = 12),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    axis.title = element_text(size = 14),
    panel.grid.major = element_line(color = "grey"),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "aliceblue"),
    strip.background = element_rect(fill = "lightgrey", color = "grey", size = 1),
    strip.text = element_text(face = "bold", size = 12, color = "navy"),
    legend.position = "right",
    legend.background = element_blank(),
    panel.margin = unit(.5, "lines"),
    panel.border = element_rect(color = "grey", fill = NA, size = 0.5)
  )
}
```

```{r echo = TRUE, message = FALSE, warning = FALSE, cache=FALSE, fig.width = 6, fig.height = 6, fig.align = "center"}
points <- rbind(points_p_day[, -1], points_p_month[, -1], points_p_year[, -1])

ggplot(points, aes(x = group, y = Freq)) + 
  geom_point(position = position_jitter(width = 0.9), alpha = 0.3) + 
  geom_boxplot(aes(color = group), size = 1, outlier.colour = NA) + 
  facet_grid(group ~ ., scales = "free") + my_theme() +
  theme(
    legend.position = "none",
    strip.placement = "outside",
    strip.background = element_blank(),
    strip.text = element_blank(),
    axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5)
  ) +
  labs(
    x = "",
    y = "Frequency",
    title = "How many data points did Google collect about me?",
    subtitle = "Number of data points per day, month and year",
    caption = "\n"
  )
```

<br>

## How accurate does Google say my data is?

Accuracy is given in meters, i.e. the smaller the better. Most data points are quite accurate.

```{r echo = TRUE, message = FALSE, warning = FALSE, cache=FALSE, fig.width = 6, fig.height = 6, fig.align = "center"}
accuracy <- data.frame(accuracy = loc$accuracy, group = ifelse(loc$accuracy < 800, "high", ifelse(loc$accuracy < 5000, "middle", "low")))

accuracy$group <- factor(accuracy$group, levels = c("high", "middle", "low"))

ggplot(accuracy, aes(x = accuracy, fill = group)) + 
  geom_histogram() + 
  facet_grid(group ~ ., scales="free") + 
  my_theme() +
  theme(
    legend.position = "none",
    strip.placement = "outside",
    strip.background = element_blank(),
    axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 0.5)
  ) +
  labs(
    x = "Accuracy",
    y = "Count",
    title = "How accurate is the location data?",
    subtitle = "Accuracy measure given by Google",
    caption = "\n"
  )
```
 
 <br>
 
# Plotting data points on maps

Finally, we are actually going to plot some maps!

The first map is a simple point plot of all locations recorded around Germany

```{r echo = TRUE, message = FALSE, warning = FALSE, cache=FALSE, fig.width = 6, fig.height = 5, fig.align = "center"}
germany <- get_map(location = 'Germany', zoom = 5)

ggmap(germany) + geom_point(data = loc, aes(x = lon, y = lat, color = accuracy), alpha = 0.1) + 
  theme(legend.position = "right") + labs(x = "Longitude", y = "Latitude", title = "Location history data points in middle Europe") +
  scale_colour_gradient(low = "red", high = "blue")
```

<br>

The second map shows a 2D bin plot of accuracy measured for all data points recorded in Münster (my home town).

```{r echo = TRUE, message = FALSE, warning = FALSE, cache=FALSE, fig.width = 6, fig.height = 5, fig.align = "center"}
munster <- get_map(location = 'Munster', zoom = 12)

options(stringsAsFactors=T)
ggmap(munster) + 
  stat_summary_2d(geom = "tile",bins = 100, data = loc, aes(x = lon, y = lat, z = accuracy), alpha=0.5) + 
  #stat_bin2d(aes(x = lon, y = lat), data = loc, bins = 200, geom = "tile") + # alternative way to plot 2d density diagram of data points
  scale_fill_gradient(low = "blue", high = "red", guide = guide_legend(title = "Accuracy")) +
  xlab("") + ylab("") + 
  theme(axis.title=element_blank(),
           axis.text=element_blank(),
           axis.ticks=element_blank()) + 
  ggtitle("Location history accuracy of Münster")
```

<br>



```{r echo = TRUE, message = FALSE, warning = FALSE, cache=FALSE, fig.width = 6, fig.height = 5, fig.align = "center"}
loc_2 <- loc[which(!is.na(loc$velocity)), ]

munster <- get_map(location = 'Munster', zoom = 10)

ggmap(munster) + geom_point(data = loc_2, aes(x = lon, y = lat, color = velocity), alpha = .1) + 
  theme(legend.position = "right") + labs(x = "Longitude", y = "Latitude", title = "Location history Münster") +
  scale_colour_gradient(low = "blue", high = "red")
```

# What distance did I travel?

Because it takes a long time to calculate, I am only looking at data from the last year.

```{r echo = TRUE, message = FALSE, warning = FALSE, cache=TRUE}
loc3 <- with(loc, subset(loc, loc$time > as.POSIXct('2016-01-01 0:00:01')))
loc3 <- with(loc, subset(loc3, loc$time < as.POSIXct('2016-10-31 23:59:59')))

# Shifting vectors for latitude and longitude to include end position
shift.vec <- function (vec, shift){
  if(length(vec) <= abs(shift)){
    rep(NA ,length(vec))
  } else {
    if (shift >= 0) {
      c(rep(NA, shift), vec[1:(length(vec)-shift)]) }
    else {
      c(vec[(abs(shift)+1):length(vec)], rep(NA, abs(shift)))
    }
  }
}

loc3$lat.p1 <- shift.vec(loc3$lat, -1)
loc3$lon.p1 <- shift.vec(loc3$lon, -1)

# Calculating distances between points (in metres) with the function pointDistance from the 'raster' package.

library(raster)
loc3$dist.to.prev <- apply(loc3, 1, FUN = function (row) {
  pointDistance(c(as.numeric(as.character(row["lat.p1"])),
                  as.numeric(as.character(row["lon.p1"]))),
                c(as.numeric(as.character(row["lat"])), as.numeric(as.character(row["lon"]))),
                lonlat = T) # Parameter 'lonlat' has to be TRUE!
})
```
```{r echo = TRUE, message = FALSE, warning = FALSE, cache=TRUE}
# distance in km
round(sum(as.numeric(as.character(loc3$dist.to.prev)), na.rm = TRUE)*0.001, digits = 2)

distance_p_month <- aggregate(loc3$dist.to.prev, by = list(month_year = as.factor(loc3$month_year)), FUN = sum)
distance_p_month$x <- distance_p_month$x*0.001
```

# Activities

Here, Google guesses my activity based on distance travelled per time.

Here again, it would take too long to look at activity from all data points.

```{r echo = TRUE, message = FALSE, warning = FALSE, cache=TRUE}
activities <- loc3$activitys

list.condition <- sapply(activities, function(x) !is.null(x[[1]]))
activities  <- activities[list.condition]

df <- do.call("rbind", activities)
main_activity <- sapply(df$activities, function(x) x[[1]][1][[1]][1])

activities_2 <- data.frame(main_activity = main_activity, 
                           time = as.POSIXct(as.numeric(df$timestampMs)/1000, origin = "1970-01-01"))

head(activities_2)
```

```{r echo = TRUE, message = FALSE, warning = FALSE, cache=FALSE, fig.width = 6, fig.height = 5, fig.align = "center"}
ggplot(activities_2, aes(x = main_activity, group = main_activity, fill = main_activity)) + geom_bar()  + my_theme()
```

<br>

-----------------

```{r echo = TRUE, message = FALSE, warning = FALSE, cache=FALSE, fig.width = 6, fig.height = 5, fig.align = "center"}
sessionInfo()
```
