---
title: "Predicting food preferences with sparklyr"
author: "Dr. Shirin Glander"
date: '`r Sys.Date()`'
output:
  prettydoc::html_pretty:
    theme: tactile
    highlight: github
---

```{r eval=FALSE}
# install a local version of Spark
library(sparklyr)
spark_install(version = "2.0.0")
```

```{r message=FALSE, warning=FALSE}
# connecting to a local Spark instance
library(sparklyr)
library(dplyr)

sc <- spark_connect(master = "local", version = "2.0.0")
```

```{r message=FALSE, warning=FALSE}
library(tidyr)
library(ggplot2)
library(ggrepel)

my_theme <- function(base_size = 12, base_family = "sans"){
  theme_minimal(base_size = base_size, base_family = base_family) +
  theme(
    axis.text = element_text(size = 12),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    axis.title = element_text(size = 14),
    panel.grid.major = element_line(color = "grey"),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "aliceblue"),
    strip.background = element_rect(fill = "lightgrey", color = "grey", size = 1),
    strip.text = element_text(face = "bold", size = 12, color = "black"),
    legend.position = "right",
    legend.justification = "top", 
    panel.border = element_rect(color = "grey", fill = NA, size = 0.5)
  )
}
```

---

The raw data behind the story "The FiveThirtyEight International Food Association's 2014 World Cup" http://fivethirtyeight.com/features/the-fivethirtyeight-international-food-associations-2014-world-cup/. For all the countries below, the response to the following question is presented: "Please rate how much you like the traditional cuisine of X"

5: I love this country's traditional cuisine. I think it's one of the best in the world.

4: I like this country's traditional cuisine. I think it's considerably above average.

3: I'm OK with this county's traditional cuisine. I think it's about average.

2: I dislike this country's traditional cuisine. I think it's considerably below average.

1: I hate this country's traditional cuisine. I think it's one of the worst in the world.

N/A: I'm unfamiliar with this country's traditional cuisine.

```{r message=FALSE, warning=FALSE}
library(fivethirtyeight)

# backup copy:
food_world_cup_copy <- food_world_cup

food_world_cup[food_world_cup == "N/A"] <- NA
food_world_cup[, 9:48][is.na(food_world_cup[, 9:48])] <- 0
food_world_cup$gender <- as.factor(food_world_cup$gender)
food_world_cup$location <- as.factor(food_world_cup$location)
```

```{r echo=FALSE, eval=FALSE, fig.width=20}
food_world_cup %>%
  gather(x, y, algeria:vietnam) %>%
  ggplot(aes(as.numeric(y))) +
    geom_density(fill = "navy", alpha = 0.7) +
    my_theme() +
  facet_wrap(~ x, ncol = 8)
```

```{r echo=FALSE, eval=FALSE, fig.width=20, fig.height=11, message=FALSE, warning=FALSE}
food_world_cup %>%
  gather(x, y, algeria:vietnam) %>%
  ggplot(aes(as.factor(y))) +
    geom_histogram(stat = "count", fill = "navy", alpha = 0.7) +
    my_theme() + 
    facet_wrap(~ x, ncol = 8)
```

```{r fig.width=15, fig.height=11, message=FALSE, warning=FALSE}
percentages <- food_world_cup %>%
  select(algeria:vietnam) %>%
  gather(x, y) %>%
  group_by(x, y) %>%
  summarise(n = n()) %>%
  mutate(Percent = round(n / sum(n) * 100, digits = 2))

percentages %>%
  ggplot(aes(x = "", y = Percent, fill = y)) + 
    geom_bar(width = 1, stat = "identity") + 
    theme_minimal() +
    coord_polar("y", start = 0) +
    facet_wrap(~ x, ncol = 8) +
    scale_fill_brewer(palette = "Set3") +
    labs(fill = "")
```


```{r eval=FALSE, message=FALSE, warning=FALSE}
library(mice)

dataset_impute <- mice(food_world_cup[, -c(1, 2)],  print = FALSE)
food_world_cup <- cbind(food_world_cup[, 2, drop = FALSE], mice::complete(dataset_impute, 1))
food_world_cup[8:47] <- lapply(food_world_cup[8:47], as.numeric)
```

```{r echo=FALSE, eval=FALSE}
save(food_world_cup, file = "food_world_cup.RData")
```

```{r echo=FALSE}
load("food_world_cup.RData")
```

```{r echo=FALSE, eval=FALSE}
countries_low_NA <- percentages %>%
  filter(y == 0) %>%
  filter(Percent < 50) %>%
  select(x)
```

```{r echo=FALSE, eval=FALSE}
food_world_cup <- food_world_cup %>%
  select(-one_of(countries_high_NA$x))
```

```{r fig.width=20}
countries <- paste(colnames(food_world_cup)[-c(1:7)])

for (response in countries) {
  food_world_cup[paste(response, "trans", sep = "_")] <- food_world_cup[response] / mean(food_world_cup[food_world_cup[response] > 0, response])
}
```

```{r eval=FALSE, echo=FALSE}
#food_world_cup <- food_world_cup %>%
  #mutate_each_(funs(. + 0.000001), countries) %>%
  #mutate_each_(funs(yjPower(., 0)), countries)
```

```{r fig.width=18, fig.height=10}
food_world_cup %>%
  gather(x, y, china_trans:vietnam_trans) %>%
  ggplot(aes(y)) +
    geom_density(fill = "navy", alpha = 0.7) +
    my_theme() + 
    facet_wrap(~ x, ncol = 8) +
    labs(x = "transformed preference")
```

### Do any countries show a gender difference?

```{r message=FALSE, warning=FALSE, fig.width=15}
food_world_cup_gather <- food_world_cup %>%
  collect %>%
  gather(country, value, china:vietnam)
                                 
food_world_cup_gather$value <- as.numeric(food_world_cup_gather$value)
food_world_cup_gather$country <- as.factor(food_world_cup_gather$country)

order <- aggregate(na.omit(food_world_cup_gather)$value, by = list(country = na.omit(food_world_cup_gather)$country), FUN = sum)

food_world_cup_gather$country <- factor(food_world_cup_gather$country, levels = order$country[order(order$x, decreasing = TRUE)])
```

```{r message=FALSE, warning=FALSE, fig.width=8}
ggplot(food_world_cup_gather, aes(x = country, y = value, fill = gender)) +
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = "Set1") +
  my_theme()
```

```{r echo=FALSE, eval=FALSE}
library(lazyeval)

for (response in colnames(food_world_cup)[-c(1:7, 23:37)]) {
  food_world_cup %>%
    collect %>%
    group_by(gender) %>%
    summarise_(mean = interp(~ mean(var), var = as.name(response))) %>%
    summarise(ratio = mean[1] / mean[2])
}
```

```{r}
food_world_cup %>%
  collect %>%
  mutate_each_(funs(as.numeric), countries) %>%
  group_by(gender) %>%
  summarise_each_(funs(mean), countries) %>%
  summarise_each_(funs(diff), countries) %>%
  gather(x, y) %>%
  ggplot(aes(x = x, y = y)) +
    geom_bar(stat = "identity", fill = "navy", alpha = 0.6) +
    my_theme() +
    labs(x = "",
         y = "difference\nbetween gender")
```

```{r}
food_world_cup <- copy_to(sc, food_world_cup, overwrite = TRUE)
```

```{r eval=FALSE, echo=FALSE, fig.width=8}
library(exprAnalysis)

heat_data <- na.omit(as.data.frame(food_world_cup))[, -c(1:7, 48:87)] %>%
  mutate_each(funs(as.numeric))
heat_data <- heat_data[which(rowSums(heat_data) > 0), ]

continent <- c("Africa", "South America", "Australia", "Europe", "Europe", "South America", "Africa", "South America", "Asia",
                         "South America", "South America", "Europe", "Central America", "South America", "Europe", "Africa", "Europe",
                         "Europe", "Africa", "Europe", "Central America", "Asia", "Asia", "Europe", "Europe", "Africa", "Asia",
                         "Central America", "Africa", "Europe", "Asia/ Europe", "Asia", "Europe", "Europe", "Asia", "Europe",
                         "Asia/ Europe", "North America", "South America", "Asia")

library(RColorBrewer)
#brewer.pal(8, "Set1")
colors <- ifelse(continent == "Africa", "#E41A1C",
                 ifelse(continent == "South America", "#377EB8",
                        ifelse(continent == "Australia", "#4DAF4A",
                               ifelse(continent == "Europe", "#984EA3",
                                      ifelse(continent == "Asia", "#FF7F00",
                                             ifelse(continent == "Central America", "#FFFF33",
                                                    ifelse(continent == "North America", "#A65628",
                                                           ifelse(continent == "Asia/ Europe", "#F781BF", NA
                                                           ))))))))


heatmaps(heat_data, method_dist = "manhattan", method_hclust = "complete", main = "", samplecols = colors)
legend("left", c("Africa", "South America", "Australia", "Europe", "Asia", "Central America", "North America", "Asia/ Europe"),
       col = brewer.pal(8, "Set1"), pch = 19, cex = 0.7)
```

```{r message=FALSE, warning=FALSE, fig.width=8, fig.height=8}
pca <- food_world_cup %>%
  mutate_each_(funs(as.numeric), countries) %>%
  ml_pca(features = paste(colnames(food_world_cup)[-c(1:47)]))

labels <- rownames(as.data.frame(pca$components))

ggplot(data = as.data.frame(pca$components), aes(x = PC1, y = PC2, color = labels, label = labels)) + 
  geom_point(size = 2, alpha = 0.6) +
  geom_text_repel() +
  labs(x = paste0("PC1: ", round(pca$explained.variance[1], digits = 2) * 100, "% variance"),
       y = paste0("PC2: ", round(pca$explained.variance[2], digits = 2) * 100, "% variance")) +
  my_theme() + 
  guides(fill = FALSE, color = FALSE)
```

```{r eval=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
library(pcaGoPromoter)

pca_func <- function(pcaOutput2, group_name){
    centroids <- aggregate(cbind(PC1, PC2) ~ groups, pcaOutput2, mean)
    conf.rgn  <- do.call(rbind, lapply(unique(pcaOutput2$groups), function(t)
          data.frame(groups = as.character(t),
                     ellipse(cov(pcaOutput2[pcaOutput2$groups == t, 1:2]),
                           centre = as.matrix(centroids[centroids$groups == t, 2:3]),
                           level = 0.95),
                     stringsAsFactors = FALSE)))
        
    plot <- ggplot(data = pcaOutput2, aes(x = PC1, y = PC2, group = groups, color = groups)) + 
      geom_polygon(data = conf.rgn, aes(fill = groups), alpha = 0.2) +
      geom_point(size = 2, alpha = 0.6) + 
      scale_color_brewer(palette = "Set1") +
      scale_fill_brewer(palette = "Set1") +
      labs(color = paste(group_name),
           fill = paste(group_name),
           x = paste0("PC1: ", round(pcaOutput$pov[1], digits = 2) * 100, "% variance"),
           y = paste0("PC2: ", round(pcaOutput$pov[2], digits = 2) * 100, "% variance")) +
      my_theme()
    
    return(plot)
}
```

```{r eval=FALSE, echo=FALSE, fig.width=5, fig.height=4}
pcaOutput <- pca(t(sapply(as.data.frame(food_world_cup)[-c(1:7, 23:37)], as.numeric)), printDropped = FALSE, scale = TRUE, center = TRUE)
pcaOutput2 <- as.data.frame(pcaOutput$scores)

pcaOutput2$groups <- as.data.frame(food_world_cup)$knowledge

pca1 <- pca_func(pcaOutput2, group_name = "knowledge")
```

```{r eval=FALSE, echo=FALSE, fig.width=5, fig.height=4}
pcaOutput2$groups <- as.data.frame(food_world_cup)$interest
pca2 <- pca_func(pcaOutput2, group_name = "interest")
```

```{r eval=FALSE, echo=FALSE, fig.width=5, fig.height=4}
pcaOutput2$groups <- as.data.frame(food_world_cup)$age
pca3 <- pca_func(pcaOutput2, group_name = "age")
```

```{r eval=FALSE, echo=FALSE, fig.width=6, fig.height=4}
pcaOutput2$groups <- as.data.frame(food_world_cup)$household_income
pca4 <- pca_func(pcaOutput2, group_name = "household_income")
```

```{r eval=FALSE, echo=FALSE, fig.width=7, fig.height=4}
pcaOutput2$groups <- as.data.frame(food_world_cup)$education
pca5 <- pca_func(pcaOutput2, group_name = "education")
```

```{r eval=FALSE, echo=FALSE, fig.width=6, fig.height=4}
pcaOutput2$groups <- as.data.frame(food_world_cup)$location
pca6 <- pca_func(pcaOutput2, group_name = "location")
```

```{r eval=FALSE, echo=FALSE, fig.width=20, fig.height=10, message=FALSE, warning=FALSE}
library(gridExtra)
library(grid)

grid.arrange(pca1, pca2, pca3, pca4, pca5, pca6, ncol = 3)
```


```{r}
food_world_cup <- tbl(sc, "food_world_cup") %>%
  ft_string_indexer(input_col = "interest", output_col = "interest_idx") %>%
  ft_string_indexer(input_col = "gender", output_col = "gender_idx") %>%
  ft_string_indexer(input_col = "age", output_col = "age_idx") %>%
  ft_string_indexer(input_col = "household_income", output_col = "household_income_idx") %>%
  ft_string_indexer(input_col = "education", output_col = "education_idx") %>%
  ft_string_indexer(input_col = "location", output_col = "location_idx") %>%
  ft_string_indexer(input_col = "knowledge", output_col = "knowledge_idx")
```

```{r echo=FALSE, eval=FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.height=9}
library(reshape2)

cor(as.data.frame(food_world_cup)[, -c(1:47, 87:94)], method = "pearson") %>%
  melt %>%
  ggplot(aes(x = Var1, y = Var2, fill = value)) + 
    geom_tile(width = 0.9, height = 0.9) +
    scale_fill_gradient2(low = "white", high = "red",  
                         limit = c(0,1), space = "Lab", 
                         name = "") +
    my_theme() +
    labs(x = "", y = "")
```

```{r echo=FALSE, eval=FALSE}
food_world_cup_NA <- food_world_cup %>%
  select(algeria:vietnam) %>%
  collect %>%
  mutate_all(funs(replace(., . == 0, NA)))

food_world_cup_NA <- copy_to(sc, food_world_cup_NA, overwrite = TRUE)
```

```{r eval=FALSE}
partitions <- food_world_cup %>%
  sdf_partition(training = 0.75, test = 0.25, seed = 753)
```

```{r eval=FALSE}
library(lazyeval)

for(response in countries) {

  features <- colnames(partitions$training)[-grep(response, colnames(partitions$training))]
  features <- features[grep("_trans|_idx", features)]

  fit <- partitions$training %>%
    filter_(interp(~ var > 0, var = as.name(response))) %>%
    ml_random_forest(intercept = FALSE, response = response, features = features, type = "classification")
  
  feature_imp <- ml_tree_feature_importance(sc, fit)
  
  features <- as.character(feature_imp[1:10, 2])
  
  fit <- partitions$training %>%
    filter_(interp(~ var > 0, var = as.name(response))) %>%
    ml_random_forest(intercept = FALSE, response = response, features = features, type = "classification")
  
  partitions$test <- partitions$test %>%
    filter_(interp(~ var > 0, var = as.name(response)))
  
  pred <- sdf_predict(fit, partitions$test) %>%
    collect
  
  pred_2 <- as.data.frame(table(pred[[response]], pred$prediction))
  pred_2$response <- response
  
  pred_sc <- select(pred, -rawPrediction, -probability)
  pred_sc <- copy_to(sc, pred_sc, overwrite = TRUE)
  
  feature_imp$response <- response
  
  f1 <- ml_classification_eval(pred_sc, response, "prediction", metric = "f1")
  wP <- ml_classification_eval(pred_sc, response, "prediction", metric = "weightedPrecision")
  wR <- ml_classification_eval(pred_sc, response, "prediction", metric = "weightedRecall")
  
  ml_eval <- data.frame(response = response,
                        f1 = f1,
                        weightedPrecision = wP,
                        weightedRecall = wR)
  
  if (response == "algeria") {
    feature_imp_df <- feature_imp
    ml_eval_df <- ml_eval
    pred_df <- pred_2
  } else {
    feature_imp_df <- rbind(feature_imp_df, feature_imp)
    ml_eval_df <- rbind(ml_eval_df, ml_eval)
    pred_df <- rbind(pred_df, pred_2)
  }
}
```

```{r echo=FALSE, eval=FALSE}
library(lazyeval)

for(response in countries) {

  features <- colnames(partitions$training)[-grep(response, colnames(partitions$training))]
  features <- features[grep("_trans|_idx", features)]

  fit <- partitions$training %>%
    ml_random_forest(intercept = FALSE, response = response, features = features, type = "classification")
  
  feature_imp <- ml_tree_feature_importance(sc, fit)
  
  features <- as.character(feature_imp[1:10, 2])
  
  fit <- partitions$training %>%
    filter_(interp(~ var > 0, var = as.name(response))) %>% # tested with and without this line of code
    ml_random_forest(intercept = FALSE, response = response, features = features, type = "classification")
  
  pred <- sdf_predict(fit, partitions$test) %>%
    collect
  
  pred_2 <- as.data.frame(table(pred[[response]], pred$prediction))
  pred_2$response <- response
  
  pred_sc <- select(pred, -rawPrediction, -probability)
  pred_sc <- copy_to(sc, pred_sc, overwrite = TRUE)
  
  feature_imp$response <- response
  
  f1 <- ml_classification_eval(pred_sc, response, "prediction", metric = "f1")
  wP <- ml_classification_eval(pred_sc, response, "prediction", metric = "weightedPrecision")
  wR <- ml_classification_eval(pred_sc, response, "prediction", metric = "weightedRecall")
  
  ml_eval <- data.frame(response = response,
                        f1 = f1,
                        weightedPrecision = wP,
                        weightedRecall = wR)
  
  if (response == "china") {
    feature_imp_df <- feature_imp
    ml_eval_df <- ml_eval
    pred_df <- pred_2
  } else {
    feature_imp_df <- rbind(feature_imp_df, feature_imp)
    ml_eval_df <- rbind(ml_eval_df, ml_eval)
    pred_df <- rbind(pred_df, pred_2)
  }
}
```

```{r echo=FALSE, eval=FALSE}
# with 0 kept in response column
# data as numeric
save(feature_imp_df, file = "feature_imp_df.RData")
save(ml_eval_df, file = "ml_eval_df.RData")
save(pred_df, file = "pred_df.RData")
```

```{r echo=FALSE, eval=FALSE}
# 0s removed from response column
# data as numeric
feature_imp_df_no_0 <- feature_imp_df
ml_eval_df_no_0 <- ml_eval_df
pred_df_no_0 <- pred_df

save(feature_imp_df_no_0, file = "feature_imp_df_no_0.RData")
save(ml_eval_df_no_0, file = "ml_eval_df_no_0.RData")
save(pred_df_no_0, file = "pred_df_no_0.RData")
```


```{r echo=FALSE}
load("feature_imp_df.RData")
load("ml_eval_df.RData")
load("pred_df.RData")

#load("feature_imp_df_no_0.RData")
#load("ml_eval_df_no_0.RData")
#load("pred_df_no_0.RData")
```

```{r fig.width=10}
order <- ml_eval_df$response[order(ml_eval_df$weightedPrecision, decreasing = TRUE)]

gather(ml_eval_df, x, y, f1:weightedRecall) %>%
  mutate(response = factor(response, levels = order)) %>%
  ggplot(aes(x = response, y = y, color = x, fill = x)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.9) +
  scale_fill_brewer(palette = "Set1") +
  my_theme() +
  labs(fill = "", color = "", x = "", y = "value")
```

```{r echo=FALSE, eval=FALSE, fig.width=10}
order <- ml_eval_df_no_0$response[order(ml_eval_df_no_0$weightedPrecision, decreasing = TRUE)]

gather(ml_eval_df_no_0, x, y, f1:weightedRecall) %>%
  mutate(response = factor(response, levels = order)) %>%
  ggplot(aes(x = response, y = y, color = x, fill = x)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_brewer(palette = "Set1") +
  my_theme()
```

```{r message=FALSE, warning=FALSE, fig.width=9, fig.height=6}
feats <- feature_imp_df %>%
  filter(response == "spain") %>%
  slice(1:10)

as.data.frame(food_world_cup) %>%
  select_(.dots = c("spain", as.character(feats$feature))) %>%
  gather(x, y, -spain) %>%
  filter(spain > 0) %>%
  ggplot(aes(x = x, y = spain, color = y)) +
  geom_jitter(alpha = 0.2) +
  scale_color_gradient(low = "blue", high = "red") +
  my_theme() +
  labs(x = "", color = "")
```

```{r message=FALSE, warning=FALSE, fig.width=9, fig.height=6}
feats <- feature_imp_df %>%
  filter(response == "greece") %>%
  slice(1:10)

as.data.frame(food_world_cup) %>%
  select_(.dots = c("greece", as.character(feats$feature))) %>%
  gather(x, y, -greece) %>%
  filter(greece > 0) %>%
  ggplot(aes(x = x, y = greece, color = y)) +
  geom_jitter(alpha = 0.2) +
  scale_color_gradient(low = "blue", high = "red") +
  my_theme() +
  labs(x = "", color = "")
```

```{r message=FALSE, warning=FALSE, fig.width=9, fig.height=6}
feats <- feature_imp_df %>%
  filter(response == "italy") %>%
  slice(1:10)

as.data.frame(food_world_cup) %>%
  select_(.dots = c("italy", as.character(feats$feature))) %>%
  gather(x, y, -italy) %>%
  filter(italy > 0) %>%
  ggplot(aes(x = x, y = italy, color = y)) +
  geom_jitter(alpha = 0.2) +
  scale_color_gradient(low = "blue", high = "red") +
  my_theme() +
  labs(x = "", color = "")
```

---

Next week, I'll be looking into H2O integration with sparklyr using rsparkling and Sparkling Water. 

------------------

<br>

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4, fig.align="center", cache=FALSE}
sessionInfo()
```
