---
title: "Predicting food preferences with sparklyr"
author: "Dr. Shirin Glander"
date: '`r Sys.Date()`'
output:
  prettydoc::html_pretty:
    theme: tactile
    highlight: github
---

Obviously, the power of Spark is in applying it to large datasets with many features (like image categorisation, etc.). Because that's not very handy for demonstration purposes, I am here showing the principles on a small dataset.

---

http://spark.rstudio.com/

http://spark.apache.org/

http://spark.apache.org/faq.html

https://bigdatauniversity.com/courses/what-is-spark/

http://spark.apache.org/docs/latest/quick-start.html

http://spark.apache.org/docs/latest/programming-guide.html


```{r eval=FALSE}
# install a local version of Spark
library(sparklyr)
spark_install(version = "2.0.0")
```

```{r message=FALSE, warning=FALSE}
# connecting to a local Spark instance
library(sparklyr)
library(dplyr)

sc <- spark_connect(master = "local", version = "2.0.0")
```

```{r message=FALSE, warning=FALSE}
library(tidyr)
library(ggplot2)
library(ggrepel)

my_theme <- function(base_size = 12, base_family = "sans"){
  theme_minimal(base_size = base_size, base_family = base_family) +
  theme(
    axis.text = element_text(size = 12),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    axis.title = element_text(size = 14),
    panel.grid.major = element_line(color = "grey"),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "aliceblue"),
    strip.background = element_rect(fill = "lightgrey", color = "grey", size = 1),
    strip.text = element_text(face = "bold", size = 12, color = "black"),
    legend.position = "right",
    legend.justification = "top", 
    panel.border = element_rect(color = "grey", fill = NA, size = 0.5)
  )
}
```

---

The raw data behind the story "The FiveThirtyEight International Food Association's 2014 World Cup" http://fivethirtyeight.com/features/the-fivethirtyeight-international-food-associations-2014-world-cup/. For all the countries below, the response to the following question is presented: "Please rate how much you like the traditional cuisine of X"

5: I love this country's traditional cuisine. I think it's one of the best in the world.

4: I like this country's traditional cuisine. I think it's considerably above average.

3: I'm OK with this county's traditional cuisine. I think it's about average.

2: I dislike this country's traditional cuisine. I think it's considerably below average.

1: I hate this country's traditional cuisine. I think it's one of the worst in the world.

N/A: I'm unfamiliar with this country's traditional cuisine.

```{r}
library(fivethirtyeight)

# backup copy:
food_world_cup_copy <- food_world_cup

food_world_cup[food_world_cup == "N/A"] <- NA
food_world_cup[, 9:48][is.na(food_world_cup[, 9:48])] <- 0
food_world_cup$gender <- as.factor(food_world_cup$gender)
food_world_cup$location <- as.factor(food_world_cup$location)
```

```{r fig.width=20}
food_world_cup %>%
  gather(x, y, algeria:vietnam) %>%
  ggplot(aes(as.numeric(y))) +
    geom_density(fill = "navy", alpha = 0.7) +
    my_theme() +
  facet_wrap(~ x, ncol = 8)
```

```{r fig.width=20}
food_world_cup %>%
  gather(x, y, algeria:vietnam) %>%
  ggplot(aes(as.factor(y))) +
    geom_histogram(stat = "count", fill = "navy", alpha = 0.7) +
    my_theme() + 
  facet_wrap(~ x, ncol = 8)
```

```{r eval=FALSE, message=FALSE, warning=FALSE}
library(mice)

dataset_impute <- mice(food_world_cup[, -c(1, 2)],  print = FALSE)
food_world_cup <- cbind(food_world_cup[, 2, drop = FALSE], mice::complete(dataset_impute, 1))
food_world_cup[8:47] <- lapply(food_world_cup[8:47], as.numeric)
```

```{r echo=FALSE, eval=FALSE}
save(food_world_cup, file = "food_world_cup.RData")
```

```{r echo=FALSE}
load("food_world_cup.RData")
```

```{r fig.width=20}
#summary(food_world_cup)

countries <- paste(colnames(food_world_cup)[-c(1:7, 48:54)])

for (response in countries) {
  food_world_cup[paste(response, "trans", sep = "_")] <- food_world_cup[response] / mean(food_world_cup[food_world_cup[response] > 0, response])
}

#food_world_cup <- food_world_cup %>%
  #mutate_each_(funs(. + 0.000001), countries) %>%
  #mutate_each_(funs(yjPower(., 0)), countries)
```

```{r fig.width=20}
food_world_cup %>%
  gather(x, y, algeria_trans:vietnam_trans) %>%
  ggplot(aes(y)) +
    geom_density(fill = "navy", alpha = 0.7) +
    my_theme() + 
    facet_wrap(~ x, ncol = 8)
```

```{r}
food_world_cup <- copy_to(sc, food_world_cup)
head(food_world_cup)
```

http://spark.rstudio.com/mllib.html

```{r message=FALSE, warning=FALSE, fig.width=15}
food_world_cup_gather <- food_world_cup %>%
  collect %>%
  gather(country, value, algeria:vietnam)
                                 
food_world_cup_gather$value <- as.numeric(food_world_cup_gather$value)
food_world_cup_gather$country <- as.factor(food_world_cup_gather$country)

order <- aggregate(na.omit(food_world_cup_gather)$value, by = list(country = na.omit(food_world_cup_gather)$country), FUN = sum)

food_world_cup_gather$country <- factor(food_world_cup_gather$country, levels = order$country[order(order$x, decreasing = TRUE)])
```

```{r message=FALSE, warning=FALSE, fig.width=8}
ggplot(food_world_cup_gather, aes(x = country, y = value, fill = gender)) +
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = "Set1") +
  my_theme()
```

### Do any countries show a gender difference?

https://cran.r-project.org/web/packages/dplyr/vignettes/nse.html

```{r echo=FALSE, eval=FALSE}
for (response in colnames(food_world_cup)[-c(1:7, 48:54)]) {
  food_world_cup %>%
    collect %>%
    group_by(gender) %>%
    summarise_(mean = interp(~ mean(var), var = as.name(response))) %>%
    summarise(ratio = mean[1] / mean[2])
}
```

```{r}
food_world_cup %>%
  collect %>%
  mutate_each_(funs(as.numeric), countries) %>%
  group_by(gender) %>%
  summarise_each_(funs(mean), countries) %>%
  summarise_each_(funs(diff), countries) %>%
  gather(x, y) %>%
  ggplot(aes(x = x, y = y)) +
    geom_bar(stat = "identity", fill = "navy", alpha = 0.6) +
    my_theme() +
    labs(x = "",
         y = "difference\nbetween gender")
```

```{r fig.width=8}
library(exprAnalysis)

heat_data <- na.omit(as.data.frame(food_world_cup))[, -c(1:7, 48:87)] %>%
  mutate_each(funs(as.numeric))
heat_data <- heat_data[which(rowSums(heat_data) > 0), ]

continent <- c("Africa", "South America", "Australia", "Europe", "Europe", "South America", "Africa", "South America", "Asia",
                         "South America", "South America", "Europe", "Central America", "South America", "Europe", "Africa", "Europe",
                         "Europe", "Africa", "Europe", "Central America", "Asia", "Asia", "Europe", "Europe", "Africa", "Asia",
                         "Central America", "Africa", "Europe", "Asia/ Europe", "Asia", "Europe", "Europe", "Asia", "Europe",
                         "Asia/ Europe", "North America", "South America", "Asia")

library(RColorBrewer)
#brewer.pal(8, "Set1")
colors <- ifelse(continent == "Africa", "#E41A1C",
                 ifelse(continent == "South America", "#377EB8",
                        ifelse(continent == "Australia", "#4DAF4A",
                               ifelse(continent == "Europe", "#984EA3",
                                      ifelse(continent == "Asia", "#FF7F00",
                                             ifelse(continent == "Central America", "#FFFF33",
                                                    ifelse(continent == "North America", "#A65628",
                                                           ifelse(continent == "Asia/ Europe", "#F781BF", NA
                                                           ))))))))


heatmaps(heat_data, method_dist = "manhattan", method_hclust = "complete", main = "", samplecols = colors)
legend("left", c("Africa", "South America", "Australia", "Europe", "Asia", "Central America", "North America", "Asia/ Europe"),
       col = brewer.pal(8, "Set1"), pch = 19, cex = 0.7)
```

```{r fig.width=8, fig.height=8}
pca <- food_world_cup %>%
  mutate_each_(funs(as.numeric), countries) %>%
  ml_pca(features = paste(colnames(food_world_cup)[-c(1:7, 48:87)]))

groups <- as.factor(colors)
labels <- rownames(as.data.frame(pca$components))

ggplot(data = as.data.frame(pca$components), aes(x = PC1, y = PC2, color = groups, label = labels)) + 
  geom_point(size = 2, alpha = 0.6) +
  geom_text_repel() +
  labs(x = paste0("PC1: ", round(pca$explained.variance[1], digits = 2) * 100, "% variance"),
       y = paste0("PC2: ", round(pca$explained.variance[2], digits = 2) * 100, "% variance")) +
  my_theme() + 
  guides(fill = FALSE, color = FALSE)
```

```{r message=FALSE, warning=FALSE}
library(pcaGoPromoter)

pca_func <- function(pcaOutput2, group_name){
    centroids <- aggregate(cbind(PC1, PC2) ~ groups, pcaOutput2, mean)
    conf.rgn  <- do.call(rbind, lapply(unique(pcaOutput2$groups), function(t)
          data.frame(groups = as.character(t),
                     ellipse(cov(pcaOutput2[pcaOutput2$groups == t, 1:2]),
                           centre = as.matrix(centroids[centroids$groups == t, 2:3]),
                           level = 0.95),
                     stringsAsFactors = FALSE)))
        
    plot <- ggplot(data = pcaOutput2, aes(x = PC1, y = PC2, group = groups, color = groups)) + 
      geom_polygon(data = conf.rgn, aes(fill = groups), alpha = 0.2) +
      geom_point(size = 2, alpha = 0.6) + 
      scale_color_brewer(palette = "Set1") +
      scale_fill_brewer(palette = "Set1") +
      labs(color = paste(group_name),
           fill = paste(group_name),
           x = paste0("PC1: ", round(pcaOutput$pov[1], digits = 2) * 100, "% variance"),
           y = paste0("PC2: ", round(pcaOutput$pov[2], digits = 2) * 100, "% variance")) +
      my_theme()
    
    return(plot)
}
```

```{r fig.width=5, fig.height=4}
pcaOutput <- pca(t(sapply(as.data.frame(food_world_cup)[-c(1:7, 48:87)], as.numeric)), printDropped = FALSE, scale = TRUE, center = TRUE)
pcaOutput2 <- as.data.frame(pcaOutput$scores)

pcaOutput2$groups <- as.data.frame(food_world_cup)$knowledge

pca1 <- pca_func(pcaOutput2, group_name = "knowledge")
```

```{r fig.width=5, fig.height=4}
pcaOutput2$groups <- as.data.frame(food_world_cup)$interest
pca2 <- pca_func(pcaOutput2, group_name = "interest")
```

```{r fig.width=5, fig.height=4}
pcaOutput2$groups <- as.data.frame(food_world_cup)$age
pca3 <- pca_func(pcaOutput2, group_name = "age")
```

```{r fig.width=6, fig.height=4}
pcaOutput2$groups <- as.data.frame(food_world_cup)$household_income
pca4 <- pca_func(pcaOutput2, group_name = "household_income")
```

```{r fig.width=7, fig.height=4}
pcaOutput2$groups <- as.data.frame(food_world_cup)$education
pca5 <- pca_func(pcaOutput2, group_name = "education")
```

```{r fig.width=6, fig.height=4}
pcaOutput2$groups <- as.data.frame(food_world_cup)$location
pca6 <- pca_func(pcaOutput2, group_name = "location")
```

```{r fig.width=20, fig.height=10, message=FALSE, warning=FALSE}
library(gridExtra)
library(grid)

grid.arrange(pca1, pca2, pca3, pca4, pca5, pca6, ncol = 3)
```

---

```{r}
food_world_cup <- tbl(sc, "food_world_cup") %>%
  ft_string_indexer(input_col = "interest", output_col = "interest_idx") %>%
  ft_string_indexer(input_col = "gender", output_col = "gender_idx") %>%
  ft_string_indexer(input_col = "age", output_col = "age_idx") %>%
  ft_string_indexer(input_col = "household_income", output_col = "household_income_idx") %>%
  ft_string_indexer(input_col = "education", output_col = "education_idx") %>%
  ft_string_indexer(input_col = "location", output_col = "location_idx") %>%
  ft_string_indexer(input_col = "knowledge", output_col = "knowledge_idx")
```

The reason why Nigeria, Cameroon, etc. had such high accuracy when we keep the 0s, is that they had a majority of 0 votes, meaning that the likelihood of correctly classifying a sample as 0 is high! When we remove 0s however, the prediction accuracy drops sharply. The main problem is that the distribution of preferences and whether a dish is known (i.e. has a value > 0) is very different between countries. Specifically, whenever a value is overrepresented, the likelihood of correctly classifying that value by chance increases and thus introduces bias to our models!

```{r}
cor()

ggplot(data = associations_table_melt, aes(x = word, y = variable, fill = value)) + 
    geom_tile(width=.9, height=.9) +
    scale_fill_gradient2(low = "white", high = "red",  
                         limit = c(0,1), space = "Lab", 
                         name="") +
    my_theme() + 
    theme(
      axis.title = element_blank(),
      legend.position = "right") +
    coord_fixed() +
    labs(title = paste0("Word pair associations > ", cor, " in ", name, "'s lines"))
  
```


```{r echo=FALSE, eval=FALSE}
food_world_cup_NA <- food_world_cup %>%
  select(algeria:vietnam) %>%
  collect %>%
  mutate_all(funs(replace(., . == 0, NA)))

food_world_cup_NA <- copy_to(sc, food_world_cup_NA, overwrite = TRUE)
```

```{r}
partitions <- food_world_cup %>%
  sdf_partition(training = 0.75, test = 0.25, seed = 753)
```

```{r eval=FALSE}
library(lazyeval)

for(response in countries) {

  features <- colnames(partitions$training)[-grep(response, colnames(partitions$training))]
  features <- features[grep("_trans|_idx", features)]

  fit <- partitions$training %>%
    filter_(interp(~ var > 0, var = as.name(response))) %>% # tested with and without this line of code
    ml_random_forest(intercept = FALSE, response = response, features = features, type = "classification")
  
  feature_imp <- ml_tree_feature_importance(sc, fit)
  
  features <- as.character(feature_imp[1:10, 2])
  
  fit <- partitions$training %>%
    filter_(interp(~ var > 0, var = as.name(response))) %>% # tested with and without this line of code
    ml_random_forest(intercept = FALSE, response = response, features = features, type = "classification")
  
  partitions$test <- partitions$test %>%
    filter_(interp(~ var > 0, var = as.name(response)))
  
  pred <- sdf_predict(fit, partitions$test) %>%
    collect
  
  pred_2 <- as.data.frame(table(pred[[response]], pred$prediction))
  pred_2$response <- response
  
  pred_sc <- select(pred, -rawPrediction, -probability)
  pred_sc <- copy_to(sc, pred_sc, overwrite = TRUE)
  
  feature_imp$response <- response
  
  f1 <- ml_classification_eval(pred_sc, response, "prediction", metric = "f1")
  wP <- ml_classification_eval(pred_sc, response, "prediction", metric = "weightedPrecision")
  wR <- ml_classification_eval(pred_sc, response, "prediction", metric = "weightedRecall")
  
  ml_eval <- data.frame(response = response,
                        f1 = f1,
                        weightedPrecision = wP,
                        weightedRecall = wR)
  
  if (response == "algeria") {
    feature_imp_df <- feature_imp
    ml_eval_df <- ml_eval
    pred_df <- pred_2
  } else {
    feature_imp_df <- rbind(feature_imp_df, feature_imp)
    ml_eval_df <- rbind(ml_eval_df, ml_eval)
    pred_df <- rbind(pred_df, pred_2)
  }
}
```

```{r echo=FALSE, eval=FALSE}
# with 0 kept in response column
# data as numeric
save(feature_imp_df, file = "feature_imp_df.RData")
save(ml_eval_df, file = "ml_eval_df.RData")
save(pred_df, file = "pred_df.RData")
```

```{r echo=FALSE, eval=FALSE}
# 0s removed from response column
# data as numeric
feature_imp_df_no_0 <- feature_imp_df
ml_eval_df_no_0 <- ml_eval_df
pred_df_no_0 <- pred_df

save(feature_imp_df_no_0, file = "feature_imp_df_no_0.RData")
save(ml_eval_df_no_0, file = "ml_eval_df_no_0.RData")
save(pred_df_no_0, file = "pred_df_no_0.RData")
```


```{r echo=FALSE}
load("feature_imp_df.RData")
load("ml_eval_df.RData")
load("pred_df.RData")

load("feature_imp_df_no_0.RData")
load("ml_eval_df_no_0.RData")
load("pred_df_no_0.RData")
```

```{r fig.width=10}
order <- ml_eval_df$response[order(ml_eval_df$weightedPrecision, decreasing = TRUE)]

gather(ml_eval_df, x, y, f1:weightedRecall) %>%
  mutate(response = factor(response, levels = order)) %>%
  ggplot(aes(x = response, y = y, color = x, fill = x)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_brewer(palette = "Set1") +
  my_theme()
```

```{r fig.width=10}
order <- ml_eval_df_no_0$response[order(ml_eval_df_no_0$weightedPrecision, decreasing = TRUE)]

gather(ml_eval_df_no_0, x, y, f1:weightedRecall) %>%
  mutate(response = factor(response, levels = order)) %>%
  ggplot(aes(x = response, y = y, color = x, fill = x)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_brewer(palette = "Set1") +
  my_theme()
```

```{r message=FALSE, warning=FALSE, fig.width=10}
as.data.frame(food_world_cup)[-c(1:7, 48:94)] %>%
  gather(x, y, -nigeria) %>%
  ggplot(aes(x = x, y = nigeria, color = y)) +
  geom_jitter(alpha = 0.2) +
  scale_color_gradient(low = "blue", high = "red") +
  my_theme()
```

```{r message=FALSE, warning=FALSE, fig.width=10}
as.data.frame(food_world_cup)[-c(1:7, 48:94)] %>%
  gather(x, y, -cameroon) %>%
  filter(y > 0) %>%
  ggplot(aes(x = x, y = cameroon, color = y)) +
  geom_jitter(alpha = 0.2) +
  scale_color_gradient(low = "blue", high = "red") +
  my_theme()
```

```{r message=FALSE, warning=FALSE, fig.width=10}
as.data.frame(food_world_cup)[-c(1:7, 48:94)] %>%
  gather(x, y, -mexico) %>%
  ggplot(aes(x = x, y = mexico, color = y)) +
  geom_jitter(alpha = 0.2) +
  scale_color_gradient(low = "blue", high = "red") +
  my_theme()
```

```{r message=FALSE, warning=FALSE, fig.width=10}
as.data.frame(food_world_cup)[-c(1:7, 48:94)] %>%
  gather(x, y, -italy) %>%
  ggplot(aes(x = x, y = italy, color = y)) +
  geom_jitter(alpha = 0.2) +
  scale_color_gradient(low = "blue", high = "red") +
  my_theme()
```

------------------

<br>

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4, fig.align="center", cache=FALSE}
sessionInfo()
```
