---
title: "Explore World Gender Statistics with Shiny"
author: "Dr. Shirin Glander"
date: '`r Sys.Date()`'
output:
  md_document:
    variant: markdown_github
---

This week I explored the World Gender Statistics dataset. You can look at 160 measurements over 56 years with [my Shiny app here](https://shiring.shinyapps.io/wgs_app/).

<br>

I prepared the data as follows:

### Data.csv

- Country.Name: the name of the country
- Country.Code: the country's code
- Indicator.Name: the name of the variable that this row represents
- Indicator.Code: a unique id for the variable
- 1960 - 2016: one column EACH for the value of the variable in each year it was available

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=8, fig.align="center", cache=TRUE}
dataset <- read.csv("Data.csv")
dataset_subs <- dataset[grep(".FE|.MA", dataset$Indicator.Code), ]
head(dataset_subs)

dataset_subs$Indicator.Name <- as.character(dataset_subs$Indicator.Name)

dataset_fem <- dataset[grep("female", dataset$Indicator.Name), ]
dataset_fem$Indicator.Name <- gsub("female", "", dataset_fem$Indicator.Name)
dataset_fem$Indicator.Code <- gsub(".FE", "", dataset_fem$Indicator.Code)
dataset_fem$gender <- "female"

dataset_male <- dataset[-grep("female", dataset$Indicator.Name), ]
dataset_male$Indicator.Name <- gsub("male", "", dataset_male$Indicator.Name)
dataset_male$Indicator.Code <- gsub(".FE", "", dataset_male$Indicator.Code)
dataset_male$gender <- "male"

dataset_fem <- dataset_fem[which(dataset_fem$Indicator.Name %in% dataset_male$Indicator.Name), ]
dataset_male <- dataset_male[which(dataset_male$Indicator.Name %in% dataset_fem$Indicator.Name), ]

dataset_fem <- dataset_fem[which(dataset_fem$Country.Code %in% dataset_male$Country.Code), ]
dataset_male <- dataset_male[which(dataset_male$Country.Code %in% dataset_fem$Country.Code), ]

library(dplyr)
dataset_fem <- arrange(dataset_fem, Country.Code)
dataset_male <- arrange(dataset_male, Country.Code)

dataset_fem$Country.Code <- as.character(dataset_fem$Country.Code)
dataset_male$Country.Code <- as.character(dataset_male$Country.Code)

save(dataset_fem, file = "dataset_fem.RData")
save(dataset_male, file = "dataset_male.RData")
```

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=8, fig.align="center", cache=TRUE}
length(unique(dataset_fem$Indicator.Name)) == length(unique(dataset_male$Indicator.Name))

for (n in 1:length(unique(dataset_fem$Indicator.Name))) {
  
  code <- unique(dataset_fem$Indicator.Name)[n]
  
  print(code)
                 
  fem <- dataset_fem[which(dataset_fem$Indicator.Name == code), ]
  male <- dataset_male[which(dataset_male$Indicator.Name == code), ]

  for (i in 1:nrow(fem)) {
    
    if (i == 1) {
      
      diff <- male[i, 5:61] / fem[i, 5:61]
      diff_table <- cbind(male[i, c(1:4)], diff)
      
    } else {
      
      diff <- male[i, 5:61] / fem[i, 5:61]
      diff_table <- rbind(diff_table, 
                          cbind(male[i, c(1:4)], diff))
      
    }
  }
  
  if (n == 1) {
    
    diff_table_bind <- diff_table
    
  } else {
    
    diff_table_bind <- rbind(diff_table_bind, diff_table)
    
  }
  
}

diff_table_bind$Country.Code <- as.character(diff_table_bind$Country.Code)
diff_table_bind[diff_table_bind == "NaN"] <- NA

save(diff_table_bind, file = "diff_table_bind.RData")
```

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=8, fig.align="center", cache=TRUE}
measures <- unique(diff_table_bind$Indicator.Name)
save(measures, file = "measures.RData")

years <- gsub("X", "", colnames(diff_table_bind)[-c(1:4)])
years <- years[-length(years)]
save(years, file = "years.RData")
```

## Map

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=8, fig.align="center", cache=TRUE}
library(plyr)
library(rgdal)

wmap_countries <- readOGR(dsn="shapefiles", layer="ne_110m_admin_0_countries")

wmap_countries_df <- fortify(wmap_countries)
wmap_countries@data$id <- rownames(wmap_countries@data)
wmap_countries_df_final <- join(wmap_countries_df, wmap_countries@data, by = "id")

wmap_countries_df_final$gu_a3 <- as.character(wmap_countries_df_final$gu_a3)

save(wmap_countries_df_final, file = "wmap_countries_df_final.RData")
```

------------------

<br>

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4, fig.align="center", cache=FALSE}
sessionInfo()
```
