---
title: "Explore World Gender Statistics with Shiny"
author: "Dr. Shirin Glander"
date: '`r Sys.Date()`'
output:
  md_document:
    variant: markdown_github
---

This week I explored the World Gender Statistics dataset. You can look at 160 measurements over 56 years with [my Shiny app here](https://shiring.shinyapps.io/wgs_app/).

<br>

I prepared the data as follows:

### Data.csv

- Country.Name: the name of the country
- Country.Code: the country's code
- Indicator.Name: the name of the variable that this row represents
- Indicator.Code: a unique id for the variable
- 1960 - 2016: one column EACH for the value of the variable in each year it was available

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=8, fig.align="center", cache=TRUE}
dataset <- read.csv("Data.csv")
dataset_subs <- dataset[grep(".FE|.MA", dataset$Indicator.Code), ]
head(dataset_subs)

dataset_subs$Indicator.Name <- as.character(dataset_subs$Indicator.Name)

dataset_fem <- dataset[grep("female", dataset$Indicator.Name), ]
dataset_fem$Indicator.Name <- gsub("female", "", dataset_fem$Indicator.Name)
dataset_fem$Indicator.Name <- gsub(",", "", dataset_fem$Indicator.Name)
dataset_fem$Indicator.Code <- gsub(".FE", "", dataset_fem$Indicator.Code)
dataset_fem$gender <- "female"

dataset_male <- dataset[-grep("female", dataset$Indicator.Name), ]
dataset_male$Indicator.Name <- gsub("male", "", dataset_male$Indicator.Name)
dataset_male$Indicator.Name <- gsub(",", "", dataset_male$Indicator.Name)
dataset_male$Indicator.Code <- gsub(".FE", "", dataset_male$Indicator.Code)
dataset_male$gender <- "male"

dataset_fem <- dataset_fem[which(dataset_fem$Indicator.Name %in% dataset_male$Indicator.Name), ]
dataset_male <- dataset_male[which(dataset_male$Indicator.Name %in% dataset_fem$Indicator.Name), ]

dataset_fem <- dataset_fem[which(dataset_fem$Country.Code %in% dataset_male$Country.Code), ]
dataset_male <- dataset_male[which(dataset_male$Country.Code %in% dataset_fem$Country.Code), ]

library(dplyr)
dataset_fem <- arrange(dataset_fem, Country.Code)
dataset_male <- arrange(dataset_male, Country.Code)

dataset_fem$Country.Code <- as.character(dataset_fem$Country.Code)
dataset_male$Country.Code <- as.character(dataset_male$Country.Code)

save(dataset_fem, file = "dataset_fem.RData")
save(dataset_male, file = "dataset_male.RData")
```

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=8, fig.align="center", cache=TRUE}
length(unique(dataset_fem$Indicator.Name)) == length(unique(dataset_male$Indicator.Name))

for (n in 1:length(unique(dataset_fem$Indicator.Name))) {
  
  code <- unique(dataset_fem$Indicator.Name)[n]
  
  print(code)
                 
  fem <- dataset_fem[which(dataset_fem$Indicator.Name == code), ]
  male <- dataset_male[which(dataset_male$Indicator.Name == code), ]

  for (i in 1:nrow(fem)) {
    
    if (i == 1) {
      
      diff <- (male[i, 5:61] + 0.001) / (fem[i, 5:61] + 0.001)
      diff_table <- cbind(male[i, c(1:4)], diff)
      
    } else {
      
      diff <- (male[i, 5:61] + 0.001) / (fem[i, 5:61] + 0.001)
      diff_table <- rbind(diff_table, 
                          cbind(male[i, c(1:4)], diff))
      
    }
  }
  
  if (n == 1) {
    
    diff_table_bind <- diff_table
    
  } else {
    
    diff_table_bind <- rbind(diff_table_bind, diff_table)
    
  }
  
}


diff_table_bind$Country.Code <- as.character(diff_table_bind$Country.Code)

```

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=8, fig.align="center", cache=TRUE}
measures <- unique(diff_table_bind$Indicator.Name)
save(measures, file = "measures.RData")

years <- gsub("X", "", colnames(diff_table_bind)[-c(1:4)])
years <- years[-length(years)]
save(years, file = "years.RData")
```

## Map

- From Natural Earth .com
- changed borders to -200m
- mercator projection

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=8, fig.align="center", cache=TRUE}
library(rgdal)
library(ggplot2)
library(plyr)
library(dplyr)
library(scales)

wmap_countries <- readOGR(dsn="shapefiles/changed_borders", layer="ne_110m_admin_0_countries_smaller_wm")

wmap_countries_df <- fortify(wmap_countries)
wmap_countries@data$id <- rownames(wmap_countries@data)
wmap_countries_df_final <- join(wmap_countries_df, wmap_countries@data, by = "id")

wmap_countries_df_final$adm0_a3 <- as.character(wmap_countries_df_final$adm0_a3)

save(wmap_countries_df_final, file = "wmap_countries_smaller_df_final.RData")

diff_table_bind <- diff_table_bind[which(diff_table_bind$Country.Code %in% wmap_countries_df_final$adm0_a3), ]

save(diff_table_bind, file = "diff_table_bind.RData")

countries <- as.character(unique(diff_table_bind$Country.Name))
save(countries, file = "countries.RData")
```

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=8, fig.align="center", cache=TRUE}
load("measures.RData")
load("years.RData")
load("countries.RData")

measure = measures[1]
year = years[56]

country = countries[1]

load("diff_table_bind.RData")
load("dataset_fem.RData")
load("dataset_male.RData")
load("wmap_countries_smaller_df_final.RData")

library(ggplot2)

map_theme <- list(theme(panel.grid.minor = element_blank(),
                        panel.grid.major = element_blank(),
                        panel.background = element_blank(),
                        plot.background = element_rect(fill = "white"),
                        panel.border = element_blank(),
                        axis.line = element_blank(),
                        axis.text.x = element_blank(),
                        axis.text.y = element_blank(),
                        axis.ticks = element_blank(),
                        axis.title.x = element_blank(),
                        axis.title.y = element_blank(),
                        plot.title = element_text(size = 18)))

my_theme <- function(base_size = 12, base_family = "sans"){
  theme_minimal(base_size = base_size, base_family = base_family) +
  theme(
    axis.text = element_text(size = 12),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    axis.title = element_text(size = 14),
    panel.grid.major = element_line(color = "grey"),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white"),
    strip.background = element_rect(fill = "royalblue", color = "grey", size = 1),
    strip.text = element_text(face = "bold", size = 12, color = "white"),
    legend.position = "right",
    panel.border = element_rect(color = "grey", fill = NA, size = 0.5)
  )
}

colfunc <- colorRampPalette(c("yellow", "red"))
```

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=4, fig.align="center", cache=TRUE}
library(dplyr)
diff_table_timeline <- rbind(dataset_fem, dataset_male) %>%
  subset(Indicator.Name == measure) %>%
  subset(Country.Code %in% wmap_countries_df_final$adm0_a3) %>%
  subset(as.character(Country.Name) == country)

library(tidyr)
diff_table_timeline_gather <- gather(diff_table_timeline, year, value, X1960:X2015)
diff_table_timeline_gather$year <- gsub("X", "", diff_table_timeline_gather$year)

ggplot(diff_table_timeline_gather, aes(x = year, y = value, color = gender, group = gender)) +
  geom_line(size = 1, alpha = 0.7) +
  geom_point(size = 2, alpha = 0.7) +
  my_theme() +
  scale_color_brewer(palette = "Set1") +
  labs(title = paste(measure),
       subtitle = paste(country),
       x = "Year",
       y = "Value",
       color = "Gender")
```

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=4, fig.align="center", cache=TRUE}
diff_table_timeline_2 <- cbind(diff_table_timeline[, c(1:4, 62)], t(apply(diff_table_timeline[, grep("^X[0-9]+$", colnames(diff_table_timeline))], 1, function(x) 
  diff(x)
  )))

diff_table_timeline_gather <- gather(diff_table_timeline_2, year, value, X1961:X2015)
diff_table_timeline_gather$year <- gsub("X", "", diff_table_timeline_gather$year)

ggplot(diff_table_timeline_gather, aes(x = year, y = value, color = gender, group = gender)) +
  geom_line(size = 1, alpha = 0.7) +
  geom_point(size = 2, alpha = 0.7) +
  my_theme() +
  scale_color_brewer(palette = "Set1") +
  labs(title = paste("Change in", measure),
       subtitle = paste(country),
       x = "Year",
       y = "Value",
       color = "Gender")
```
```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=8, fig.align="center", cache=TRUE}
diff_table_map <- diff_table_bind[which(diff_table_bind$Indicator.Name == measure),
                                        c(1:3, which(colnames(diff_table_bind) == paste0("X", year)))]
      colnames(diff_table_map)[ncol(diff_table_map)] <- "value"

      library(dplyr)
      map <- left_join(subset(wmap_countries_df_final, !continent == "Antarctica"), diff_table_map, by = c("adm0_a3" = "Country.Code"))

      ggplot(map, aes(long, lat, group = group, fill = log2(value))) +
        coord_equal() +
        map_theme +
        geom_polygon() +
        geom_path(color = "white", size = 0.5) +
        labs(title = paste(measure),
             fill = "log2 of male / female") +
        scale_fill_gradient2(low = "blue", midpoint = 0, mid = "yellow", high = "red")
```

---

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=8, fig.align="center", cache=TRUE}
year_table <- diff_table_bind[which(diff_table_bind$Country.Code %in% wmap_countries_df_final$adm0_a3), ]

last_val <- apply(year_table[, grep("^X[0-9]+$", colnames(year_table))], 1, function(x) 
  na.omit(x)[length(na.omit(x))]
  )
last_val_df <- data.frame(unlist(last_val))
last_val_df$year_of_val <- gsub("([0-9]+)(.X)([0-9]+)", "\\3", rownames(last_val_df))
rownames(last_val_df) <- gsub("([0-9]+)(.X)([0-9]+)", "\\1", rownames(last_val_df))

year_table_last_val <- cbind(year_table[rownames(last_val_df), 1:3], last_val_df)
```

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=8, fig.align="center", cache=TRUE}
dif_sum <- apply(year_table[, grep("^X[0-9]+$", colnames(year_table))], 1, function(x) 
  sum(diff(log2(x[!is.na(x)])))
  )

dif_sum_df <- data.frame(unlist(dif_sum))

year_table_dif_sum <- cbind(year_table[rownames(dif_sum_df), 1:3], dif_sum_df)
```

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=8, fig.align="center", cache=TRUE}
map <- left_join(subset(wmap_countries_df_final, !continent == "Antarctica"), subset(year_table_last_val, Indicator.Name == measure), by = c("adm0_a3" = "Country.Code"))

      ggplot(map, aes(long, lat, group = group, fill = log2(unlist.last_val.), color = year_of_val)) +
        coord_equal() +
        map_theme +
        geom_polygon() +
        geom_path(size = 1) +
        labs(title = paste(measure),
             fill = "log2 of male / female") +
        scale_fill_gradient2(low = "blue", midpoint = 0, mid = "yellow", high = "red", na.value = "grey") +
        scale_color_brewer(palette="Set1")
```

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=8, fig.align="center", cache=TRUE}
na.countries <- unique(map[is.na(map$unlist.last_val.), "adm0_a3"])

year_table_dif_sum_na <- year_table_dif_sum
year_table_dif_sum_na$unlist.dif_sum. <- ifelse(year_table_dif_sum_na$Country.Code %in% na.countries, NA, year_table_dif_sum_na$unlist.dif_sum.)

map_2 <- left_join(subset(wmap_countries_df_final, !continent == "Antarctica"), subset(year_table_dif_sum_na, Indicator.Name == measure), by = c("adm0_a3" = "Country.Code"))

      ggplot(map_2, aes(long, lat, group = group, fill = unlist.dif_sum.)) +
        coord_equal() +
        map_theme +
        geom_polygon() +
        geom_path(size = 1, color = "white") +
        labs(title = paste(measure),
             fill = "sum of\ndifferences\nover time") +
        scale_fill_gradient2(low = "blue", midpoint = 0, mid = "yellow", high = "red", na.value = "grey")
```

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=8, fig.align="center", cache=TRUE}
subs <- map[!duplicated(map$adm0_a3), c("adm0_a3", "name", "economy", "income_grp", "unlist.last_val.")]
arrange(subset(subs, !is.na(unlist.last_val.)), economy, income_grp)
arrange(subset(subs, !is.na(unlist.last_val.)), income_grp, economy)
```

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=8, fig.align="center", cache=TRUE}
ggplot(map, aes(long, lat, group = group, fill = economy)) +
        coord_equal() +
        map_theme +
        geom_polygon() +
        geom_path(color = "white", size = 1) +
        labs(title = paste(measure),
             fill = "Economy")
```
     
```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=8, fig.align="center", cache=TRUE}
ggplot(map, aes(long, lat, group = group, fill = income_grp)) +
        coord_equal() +
        map_theme +
        geom_polygon() +
        geom_path(color = "white", size = 1) +
        labs(title = paste(measure),
             fill = "Income group")
```

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=8, fig.align="center", cache=TRUE}
stats_test <- rbind(dataset_fem, dataset_male) %>%
  subset(Indicator.Name == measure) %>%
  subset(Country.Code %in% wmap_countries_df_final$adm0_a3)

country.inf <- subset(wmap_countries_df_final[, c(17, 25, 26, 31, 32, 42, 43, 47, 48, 62)], !continent == "Antarctica")
country.inf <- country.inf[!duplicated(country.inf), ]

stats_test <- left_join(stats_test, country.inf, by = c("Country.Code" = "adm0_a3"))
stats_test <- left_join(stats_test, subset(year_table_last_val, Indicator.Name == measure), by = "Country.Code")

ggplot(stats_test, aes(x = log2(unlist.last_val.), fill = income_grp, color = income_grp)) +
  my_theme() +
  geom_density(alpha=0.3) +
  facet_grid(~ gender)
```

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=8, fig.align="center", cache=TRUE}
ggplot(stats_test, aes(x = log2(unlist.last_val.), fill = economy, color = economy)) +
  my_theme() +
  geom_density(alpha=0.3) +
  facet_grid(~ gender)
```

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=8, fig.align="center", cache=TRUE}
ggplot(stats_test, aes(x = economy, y = log2(unlist.last_val.), fill = income_grp, color = income_grp)) +
  geom_boxplot(alpha = 0.7, size = 2) +
  my_theme() +
  coord_flip() +
  scale_color_hue(l=20) +
  labs(title = paste(measure),
       y = "log2 of male / female",
       x = "Economy",
       color = "Income group",
       fill = "Income group")
```

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=8, fig.align="center", cache=TRUE}
ggplot(stats_test, aes(x = income_grp, y = log2(unlist.last_val.), fill = economy, color = economy)) +
  geom_boxplot(alpha = 0.7, size = 2) +
  my_theme() +
  coord_flip() +
  scale_color_hue(l=60) +
  labs(title = paste(measure),
       y = "log2 of male / female",
       x = "Income group",
       color = "Economy",
       fill = "Economy")
```


```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=8, fig.align="center", cache=TRUE}
kruskal.test(X2015 ~ economy, data = stats_test)
kruskal.test(X2015 ~ income_grp, data = stats_test)


```

- https://cran.r-project.org/web/packages/ordinal/vignettes/clm_intro.pdf
- https://journal.r-project.org/archive/2016-1/feys.pdf

- nonparametric stats
- pca

------------------

<br>

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4, fig.align="center", cache=FALSE}
sessionInfo()
```
