---
title: "DIY your own R blog with Github pages and Jekyll Boostrap"
author: "Shirin Glander"
date: '`r Sys.Date()`'
output: html_document
---

### This post is in reply to a request: How did I set up this R blog?

I have wanted to have my own R blog for a while before I went ahead and realised this page. I had seen all the cool things people did (and still do) with R by following [R-bloggers](https://www.r-bloggers.com/) and reading their newsletter every day! And while I am using R every day at work, the thematic scope there is very bioinformatics-centric (with a little bit of Machine Learning and lots of visualization and statistics), I wanted to have an incentive for me to regularly explore other types of analyses and other types of data that I don't normally work with.

I decided to host my blog on Github pages because I was already used to working with Github (it's where I store my [exprAnalysis R package](https://shiring.github.io/rna-seq/microarray/2016/09/28/exprAnalysis)).

<br>

# Setting up Github pages

Setting up your own Github page is free and very easy! All you need is a Github account. There, you have to create a new repository called "*username*.github.io"" (substitute *username* for your Github user name, in my case: ShirinG.github.io). I set up the rest via the terminal interface:

- Go to the folder where your repository should live (in this example called "*Test*")
- Clone the repository and change to its directory

```{r, engine = 'bash', eval = FALSE}
cd Home/Documents/Test
git clone https://github.com/username/username.github.io
cd username.github.io
git pull origin master
```

<br>

# JekyllBoostrap

Github pages also supports [Jekyll](https://jekyllrb.com/docs/home/), an easy to use static site generator perfect for blogging. I set up my blog with [JekyllBoostrap](http://jekyllbootstrap.com/), a full blog scaffold for Jekyll based blogs. Follow their [Quick Start Guide](http://jekyllbootstrap.com/usage/jekyll-quick-start.html) for instructions on how to install it.

It also gives detailed instructions on how to set up comments, analytics and permalinks.

<br>

## Customizing your page

Most of the heavy-lifting has already been done for you with JekyllBootstrap but of course there are many options for customizing and adapting your blog's layout to your liking.

The main file for customizing your page is Jekyll's "*_config.yml*" file (check back with [JekyllBoostrap](http://jekyllbootstrap.com/) for details). If you know a little bit about JavaScript, CSS and HTML, you can also directly modify the style sheets in the directory "*_includes*".

<br>

# Preparing R blog posts with RMarkdown

I am preparing all my blog posts in RStudio as RMarkdowns to easily test my code and see its output.

[This is how an example Rmarkdown document could look like](https://github.com/ShirinG/blog_posts_prep/blob/master/character_network/gilmore_girls.Rmd).

Your Rmarkdown will need a header:

```{r, eval = FALSE}
---
title: "This is a test post for my R blog"
author: "Dr. Shirin Glander"
date: '`r Sys.Date()`'
output: html_document
---
```

Below the header you can [write text, insert pictures, create tables and much more](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf).
Code chunks will go in between brackets.

```{r echo = FALSE, comment=NA}
cat("```{r, echo=TRUE}\nsummary(cars)\n```")
```

The final code chunk will look something like [this](https://blog.rstudio.org/2016/11/14/ggplot2-2-2-0/):

```{r echo = TRUE, message = FALSE, warning = FALSE, cache=FALSE, fig.width = 6, fig.height = 5, fig.align = "center"}
library(ggplot2)
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE, method = "loess") +
  labs(
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Two seaters (sports cars) are an exception because of their light weight",
    caption = "Data from fueleconomy.gov"
  )
```


Once I'm happy with my finished blog post (which up to this point is set to produce an html output), I am changing the output in the YAML header to 

```{r, eval = FALSE}
output: 
  md_document:
    variant: markdown_github
```

and knit again.
This will produce a markdown document (ending in .md).

JekyllBootstrap will need a header in the markdown document, so I am adding this manually to the top of my finished markdown. An example markdown header looks like this:

```{r, eval = FALSE}
---
layout: post
title: "This is a test post for my R blog"
date: 2016-12-01
categories: rblogging
tags: test ggplot2
---
```

It needs to contain the layout definition for post, the title and date. Categories and tags are optional but I like using them to get a nice thematic categorization of the posts I have written.

<br>

# Publishing a blog post via the terminal window

I am working in a different directory to produce my posts and test out things (in this example called "*blog_prep*"), so once I have my markdown file ready (in this example called "*example_post.md*"), I am going back to the terminal window and copy it to my blog repository's folder "*_post*" and rename it so that it has the following format: Date-name.md.

```{r, engine = 'bash', eval = FALSE}
cp ../blog_prep/example_post.md _posts
mv _posts/example_post.md _posts/2016-11-20-example_post.md
```

If you also had figures or graphs in your blog post, there will be a folder generated along with your markdown with the name "*example_post_files*". You need to copy this folder to your repository as well. If you gave your post a category, it will look for the files in a folder with the same name as your category with subfolders for the date:

```{r, engine = 'bash', eval = FALSE}
mkdir -p rblogging/2016/12/01
cp ../blog_prep/example_post_files rblogging/2016/12/01
```

<br>

## Testing your blog page locally

Before I push my updated blog to Github where it will go live, I am testing the output locally:

```{r, engine = 'bash', eval = FALSE}
bundle exec jekyll serve
```

In order to be able to run Jekyll locally, you need to follow [these excellent guidelines](http://jekyllrb.com/docs/installation/).

<br>

If my new blog post looks okay and I don't want to change anything any more, I can finally stage, commit and push the updated blog to Github.

```{r, engine = 'bash', eval = FALSE}
git add *
git commit -m "name"
git push origin master
```

<br>

# Publishing a blog post via RStudio

Alternatively, if you'd rather use RStudio for pushing changes to Github, you can also [connect your blog repository to RStudio](https://www.r-bloggers.com/rstudio-and-github/) and stage, commit and push from there.

<br>

And voilà, your new R blog post is out there for the world to see!
