---
title: "World Gender Statistics"
author: "Dr. Shirin Glander"
date: '`r Sys.Date()`'
output:
  prettydoc::html_pretty:
    theme: tactile
    highlight: github
---

https://www.kaggle.com/theworldbank/world-gender-statistics

This dataset was downloaded from The World Bank's Open Data project: http://data.worldbank.org/

### Data.csv

- Country.Name: the name of the country
- Country.Code: the country's code
- Indicator.Name: the name of the variable that this row represents
- Indicator.Code: a unique id for the variable
- 1960 - 2016: one column EACH for the value of the variable in each year it was available

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=8, fig.align="center", cache=TRUE}
dataset <- read.csv("Data.csv")
dataset_subs <- dataset[grep(".FE|.MA", dataset$Indicator.Code), ]
head(dataset_subs)

dataset_subs$Indicator.Name <- as.character(dataset_subs$Indicator.Name)

dataset_fem <- dataset[grep("female", dataset$Indicator.Name), ]
dataset_fem$Indicator.Name <- gsub("female", "", dataset_fem$Indicator.Name)
dataset_fem$Indicator.Code <- gsub(".FE", "", dataset_fem$Indicator.Code)
dataset_fem$gender <- "female"

dataset_male <- dataset[-grep("female", dataset$Indicator.Name), ]
dataset_male$Indicator.Name <- gsub("male", "", dataset_male$Indicator.Name)
dataset_male$Indicator.Code <- gsub(".FE", "", dataset_male$Indicator.Code)
dataset_male$gender <- "male"

dataset_fem <- dataset_fem[which(dataset_fem$Indicator.Name %in% dataset_male$Indicator.Name), ]
dataset_male <- dataset_male[which(dataset_male$Indicator.Name %in% dataset_fem$Indicator.Name), ]

dataset_fem <- dataset_fem[which(dataset_fem$Country.Code %in% dataset_male$Country.Code), ]
dataset_male <- dataset_male[which(dataset_male$Country.Code %in% dataset_fem$Country.Code), ]

library(dplyr)
dataset_fem <- arrange(dataset_fem, Country.Code)
dataset_male <- arrange(dataset_male, Country.Code)
```

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=8, fig.align="center", cache=TRUE}
length(unique(dataset_fem$Indicator.Name)) == length(unique(dataset_male$Indicator.Name))

for (n in 1:length(unique(dataset_fem$Indicator.Name))) {
  
  code <- unique(dataset_fem$Indicator.Name)[n]
  
  print(code)
                 
  fem <- dataset_fem[which(dataset_fem$Indicator.Name == code), ]
  male <- dataset_male[which(dataset_male$Indicator.Name == code), ]

  for (i in 1:nrow(fem)) {
    
    if (i == 1) {
      
      diff <- male[i, 5:61] / fem[i, 5:61]
      diff_table <- cbind(male[i, c(1:4)], diff)
      
    } else {
      
      diff <- male[i, 5:61] / fem[i, 5:61]
      diff_table <- rbind(diff_table, 
                          cbind(male[i, c(1:4)], diff))
      
    }
  }
  
  if (n == 1) {
    
    diff_table_bind <- diff_table
    
  } else {
    
    diff_table_bind <- rbind(diff_table_bind, diff_table)
    
  }
  
}

str(diff_table_bind)

diff_table_bind$Country.Code <- as.character(diff_table_bind$Country.Code)

save(diff_table_bind, file = "diff_table_bind.RData")
```

## Map

https://www.mirosa.org/blog/post/2015/01/maps-in-r/

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=8, fig.align="center", cache=TRUE}
library(plyr)
library(rgdal)

wmap_countries <- readOGR(dsn="shapefiles", layer="ne_110m_admin_0_countries")
wmap_countries_robin <- spTransform(wmap_countries, CRS("+proj=robin"))
wmap_countries_robin_df <- fortify(wmap_countries_robin)
wmap_countries_robin@data$id <- rownames(wmap_countries_robin@data)
wmap_countries_robin_df_final <- join(wmap_countries_robin_df, wmap_countries_robin@data, by="id")

str(wmap_countries_robin_df_final)

wmap_countries_robin_df_final$gu_a3 <- as.character(wmap_countries_robin_df_final$gu_a3)

save(wmap_countries_robin_df_final, file = "wmap_countries_robin_df_final.RData")
```

```{r echo=TRUE, message=FALSE, warning=FALSE, fig.width=16, fig.height=8, fig.align="center", cache=TRUE}
load("diff_table_bind.RData")
load("wmap_countries_robin_df_final.RData")

head(diff_table_bind)
head(wmap_countries_robin_df_final)

measures <- unique(diff_table_bind$Indicator.Name)
save(measures, file = "measures.RData")

years <- gsub("X", "", colnames(diff_table_bind)[-c(1:4)])
save(years, file = "years.RData")
```

```{r echo=TRUE, message=FALSE, warning=FALSE, fig.width=10, fig.height=5, fig.align="center", cache=TRUE}
library(dplyr)
library(ggplot2)
library(pals)

map_theme <- list(theme(panel.grid.minor = element_blank(),
                         panel.grid.major = element_blank(),
                         panel.background = element_blank(),
                         plot.background = element_rect(fill="white"),
                         panel.border = element_blank(),
                         axis.line = element_blank(),
                         axis.text.x = element_blank(),
                         axis.text.y = element_blank(),
                         axis.ticks = element_blank(),
                         axis.title.x = element_blank(),
                         axis.title.y = element_blank(),
                         plot.title = element_text(size=18)))

colfunc <- colorRampPalette(c("pink", "red"))
```

```{r echo=TRUE, message=FALSE, warning=FALSE, fig.width=10, fig.height=5, fig.align="center", cache=TRUE}
measure = measures[1]
year = years[56]

diff_table_map <- diff_table_bind[which(diff_table_bind$Indicator.Name == measure), c(1:3, which(colnames(diff_table_bind) == paste0("X", year)))]
colnames(diff_table_map)[ncol(diff_table_map)] <- "value"

map <- left_join(subset(wmap_countries_robin_df_final, !continent == "Antarctica"), diff_table_map, by = c("gu_a3" = "Country.Code"))

ggplot(map) +
      coord_equal() +
      map_theme +
      geom_polygon(aes(long, lat, group=group, fill=log2(value))) +
      geom_path(aes(long, lat, group=group), color="white", size=0.5) +
      labs(title = paste(measure)) +
      scale_fill_gradientn(colours=colfunc(100), guide = "colourbar")


fem_table_map <- dataset_fem[which(dataset_fem$Indicator.Name == measure), c(1:3, which(colnames(dataset_fem) == paste0("X", year)))]
colnames(fem_table_map)[ncol(fem_table_map)] <- "value"

map_fem <- left_join(subset(wmap_countries_robin_df_final, !continent == "Antarctica"), fem_table_map, by = c("gu_a3" = "Country.Code"))

ggplot(map_fem) +
      coord_equal() +
      map_theme +
      geom_polygon(aes(long, lat, group=group, fill=value)) +
      geom_path(aes(long, lat, group=group), color="white", size=0.5) +
      labs(title = paste(measure),
           subtitle = "female") +
      scale_fill_gradientn(colours=colfunc(100), guide = "colourbar")


male_table_map <- dataset_male[which(dataset_male$Indicator.Name == measure), c(1:3, which(colnames(dataset_male) == paste0("X", year)))]
colnames(male_table_map)[ncol(male_table_map)] <- "value"

map_male <- left_join(subset(wmap_countries_robin_df_final, !continent == "Antarctica"), male_table_map, by = c("gu_a3" = "Country.Code"))

ggplot(map_male) +
      coord_equal() +
      map_theme +
      geom_polygon(aes(long, lat, group=group, fill=value)) +
      geom_path(aes(long, lat, group=group), color="white", size=0.5) +
      labs(title = paste(measure),
           subtitle = "male") +
      scale_fill_gradientn(colours=colfunc(100), guide = "colourbar")
```

- individual plots for males and females
- animated plot with years

------------------

<br>

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4, fig.align="center", cache=FALSE}
sessionInfo()
```
